### 2023-05-15 Rmarkdown/paper-si.Rmd
first commit

---
title: "*Supplementary information for:* Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(echo = FALSE, dev = "png", dpi = 600)
options(dplyr.summarise.inform = FALSE)
source(here::here("R/03-inform-prior-ml.R"))
source(here::here("R/04-model-ml.R"))
library("posterior")
wq <- read_csv(here::here("data-clean/summary-table.csv"))
wq_si <- read_csv(here::here("data-clean/summary-table-si.csv"))
```

```{r functions}
# this is the opposite of fill()

unfill_vec <- function(x, placeholder = "") {
  same <- x == dplyr::lag(x)
  ifelse(!is.na(same) & same, placeholder, as.character(x))
}

```

This document has 5 pages, 5 figures, and 1 table.

```{r summary-wq-si}
options(knitr.kable.NA = "")
wq_si %>% 
  transmute(
    Parameter = unfill_vec(param),
    Unit = unfill_vec(str_replace(unit, "ug", "µg")),
    Zone = zone,
    Median = round(med, 1), 
    `Lower quartile` = round(lq, 1),
    `Upper quartile` = round(uq, 1)
  ) %>% 
  knitr::kable(caption = "**Table S1.** Summary of water quality in pipe rack effluent, by zone.")

```

```{r figure-s1, fig.cap="**Figure S1.** An example of the pipe racks installed in Zones 1 and 2."}
knitr::include_graphics(here::here("figures/figure-s1.jpg"))
```

```{r figure-s2, fig.cap="**Figure S2.** An example instruction sheet distributed to volunteer residents collecting point-of-use samples from sentinel homes."}
knitr::include_graphics(here::here("figures/figure-s2.png"))
```

```{r figure-s3, fig.cap="**Figure S3. (a)** In the full (Zones 1 and 2) model, local multi-year smooths capturing the deviations of each series from the global and location-specific smooths. **(b)** Coagulation pH at the treatment plant supplying Zone 2."}
knitr::include_graphics(here::here("figures/figure-s3.png"))
```

```{r figure-s4, fig.cap="**Figure S4.** In the Zone 1 model, local multi-year smooths capturing the deviations of each series from the global and seasonal smooths."}
knitr::include_graphics(here::here("figures/figure-s4.png"))
```

```{r figure-s5, fig.cap="**Figure S5.** Total aluminum in pipe rack effluent."}
knitr::include_graphics(here::here("figures/figure-s5.png"))
```

### 2023-05-15 Rmarkdown/paper.Rmd
first commit

---
title: "Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"
author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
output:
  word_document:
    reference_docx: StylesTemplate.docx
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(echo = FALSE, dev = "png", dpi = 600)
options(dplyr.summarise.inform = FALSE)
source(here::here("R/03-inform-prior-ml.R"))
source(here::here("R/04-model-ml.R"))
library("posterior")
wq <- read_csv(here::here("data-clean/summary-table.csv"))
wq_si <- read_csv(here::here("data-clean/summary-table-si.csv"))
```

```{r functions}
# this is the opposite of fill()

unfill_vec <- function(x, placeholder = "") {
  same <- x == dplyr::lag(x)
  ifelse(!is.na(same) & same, placeholder, as.character(x))
}

```

```{r diff-estimate}
# difference estimates for text:
summary_fun <- function(x, .fn = \(y) round(100 * exp(y))) {
  x %>%
    summarize_draws(~ quantile(.x, probs = c(0.025, .5, 0.975))) %>% 
    mutate(across(where(is.double), .fn)) %>% 
    rename_with(~ paste0("d_", str_remove(.x, "%")), .cols = where(is.numeric))
}

d <- as_draws_df(model, "b_round2_after") %>% 
  summary_fun()
dnp <- as_draws_df(model_noprior, "b_round2_after") %>% 
  summary_fun()
```

# Abstract

Orthophosphate is used widely to control lead release from plumbing into tap water. Its effect can be difficult to quantify, though, since tap water lead concentrations are site-specific. Sentinel homes with lead service lines are ideal for evaluating orthophosphate corrosion control programs, but best practices dictate the removal of lead service lines once they are identified. Sentinel homes, then, often have too short a useful life to be used effectively. Here we explore an alternative: sentinel pipe racks constructed with recovered lead pipes and supplied with water directly from the distribution system. We also propose a strategy for analyzing pipe rack data based on the generalized additive model, which approximates time series as a sum of smooth functions. In this study, geometric mean lead release from pipe racks exhibited a pronounced dose response, falling by `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%) after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_increase, c(ymin, ymax)))))), collapse = "--")`%) after a decrease to 1.5 mg PO~4~ L^-1^. Data from the sentinel homes were largely consistent with those from pipe racks: geometric mean lead levels at the high orthophosphate dose (2 mg L^-1^) were `r d$d_50`% of those at the low dose (1 mg L^-1^, 95% credible interval: `r d$d_2.5`--`r d$d_97.5`%). Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring. They also provide a fully Bayesian framework for quantifying orthophosphate's effect on lead release that is well-suited to incorporating information from multiple sources.

*Keywords:* EPA LCRR; Health Canada; corrosion control; generalized additive model; Bayesian multilevel model

```{r toc-art}
knitr::include_graphics(here::here("figures/figure-toc.png"))
```

# Introduction

Updated regulations on lead in drinking water promise to expedite replacement of lead service lines in Canada and the USA. Even afterwards, though, a substantial legacy of lead plumbing---including lead:tin solder and brass---will have to be managed. This will require careful control of drinking water chemistry to limit lead solubility and maintain durable corrosion scale.

Orthophosphate is an important tool to that end [@schock_understanding_1989; @dodrill_corrosion_1995; @bae_impact_2020; @dore_study_2019]. It works by forming low-solubility lead-phosphate minerals like pyromorphite (Pb~5~(PO~4~)~3~(Cl,F,OH)) [@lytle_lead_2020] and phosphohedyphane (Ca~2~Pb~3~(PO~4~)~3~(Cl,F,OH) [@bae_ability_2020]. Sometimes, it can be effective without forming a lead-phosphate phase [@aghasadeghi_pilotscale_2021], perhaps by blocking active sites on lead carbonate surfaces [@noel_effect_2014; @li_impact_2021] or by forming an amorphous diffusion barrier with iron, aluminum, manganese, or calcium [@wasserstrom_scale_2017; @tully_water_2019].

It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics. And while lead solubility modeling can be informative, it fails to account for the complex mineralogy of lead corrosion scale or the generation of particles [@tully_water_2019; @trueman_characterizing_2019]. A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022]. But to protect the inhabitants' health, lead pipe is often replaced once identified. Sentinel homes, then, may have too short a life to be useful in monitoring plumbosolvency changes.

Here we describe an alternative: sentinel lead pipe racks operated with feedwater directly from the distribution system. While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible. Sentinel pipe racks can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind. And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022].

We present data from three separate racks, located at three sites within the Halifax Regional Municipality, a medium-sized Canadian city. We used a robust hierarchical Bayesian generalized additive model with continuous-time autoregressive errors [@trueman_comparing_2022] to estimate the effect on lead release of a dose increase from 1 to 2 mg PO~4~ L^-1^. Then, we used this estimate as a prior probability for the same effect in nine sentinel homes. Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^. Our results provide a fully Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.

# Materials and methods

Data were collected in a single water system with two zones supplied by different source waters and treatment plants. Zone 1 is supplied by a conventional treatment plant employing alum coagulation, flocculation, clarification, and filtration. Zone 2 is supplied by a plant employing alum coagulation, flocculation, and direct filtration. Across the two zones, thousands of lead service lines remain, all of which will be replaced by 2038 as a part of the utility's comprehensive replacement program [@krkosek_halifax_2022].

## Water quality

Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with a pooled median pH and dissolved inorganic carbon concentration in pipe rack effluent of 7.3 and 4.4 mg C L^-1^ (Table 1). And while water quality in Zones 1 and 2 was largely similar, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022]. Aluminum concentrations in Zone 1 were much lower and more consistent throughout the year (Table S1).

```{r summary-wq}
options(knitr.kable.NA = "")
wq %>% 
  transmute(
    Parameter = param,
    Unit = str_replace(unit, "ug", "µg"),
    Median = round(med, 1), 
    `Lower quartile` = round(lq, 1),
    `Upper quartile` = round(uq, 1)
  ) %>% 
  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent; these pooled estimates represent both zones (zone-specific water quality is summarized in Table S1).")
```

## Data collection

### Sentinel pipe racks

Pipe racks were installed in utility-owned infrastructure; two were located in Zone 1 and one in Zone 2. Each was fitted with four replicate recovered lead pipe sections, supplied in parallel with water from the distribution system (an example is shown in Figure S1). Each pipe was excavated and handled according to principles outlined in a recent paper [@harmon_holistic_2022] and was approximately 60 cm long with an internal diameter of 1.3 cm. Each was connected to plastic tubing at either end with a brass compression fitting, yielding two galvanic lead-brass connections per pipe. A timed valve supplied flow to the pipe sections for two minutes every six hours, and samples were collected approximately monthly, as the valves opened, at a nominal flow rate of 8 L min^-1^.

### Sentinel homes

Of the nine sentinel homes, seven were supplied by partial lead service lines (private lead, public copper) and the remaining two by copper service lines; all were located in Zone 2. At each sampling round, volunteer residents collected four consecutive 1L samples, starting with the first-draw after a minimum six-hour stagnation period. This 4 $\times$ 1L profile was followed first by a 10-minute flush of the plumbing and then by collection of a final 1L sample.  Sample profiles were collected in May--June 2021, at 1 mg PO~4~ L^-1^, and again in May--June 2022, at 2 mg PO~4~ L^-1^. An example instruction sheet provided to residents is included as Figure S2. During the study, all residents were provided with pitcher filters certified by NSF for removal of lead.

### Analytical methods

An accredited laboratory measured lead, iron, manganese, zinc, and aluminum [@epa2014method], as well as dissolved inorganic and total organic carbon [@sm_5310], chloride [@sm_4500_Cl], sulfate [@astm_516], orthophosphate [@sm_4500_P], and alkalinity [@epa1974method] in pipe rack effluent samples. Turbidity, pH, free chlorine, temperature, conductivity, dissolved oxygen, and oxygen reduction potential were determined onsite using portable Hach instruments. Orthophosphate was also quantified [@sm_4500_P] in treated water by Zone 1 and 2 treatment plant staff.

## Data analysis

We used R [@R], and a collection of contributed packages [@rmarkdown; @tidyverse; @readxl; @janitor; @assertr; @brms; @ggdist; @glue; @posterior; @testthat; @bayesplot; @patchwork; @ggtext; @devtools; @mgcv], to analyze and visualize the data. The code and data necessary to reproduce the main results of the paper are available online [@cct_monitoring].

### Sentinel pipe racks

Lead in pipe rack effluent, $y_t$, was modeled using a robust hierarchical Bayesian generalized additive model (GAM) with continuous-time first-order autoregressive errors [@trueman_comparing_2022; @bgamcar1; @pedersen_hierarchical_2019; @simpson_modelling_2018]. The model is specified in equation (1),

$$
(1)~~
\begin{align}
\text{likelihood:} \\
log(y_t) \sim T(\mu_t,\sigma,\nu) \\ 
\\
\text{model for }\mu_t\text{:} \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
f_j(t) = X_j\beta_j + Z_jb_j \\
r_{t-s} = log(y_{t-s}) - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s) \\
\\
\text{priors:} \\
\sigma \sim Half \text{-} T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.1) \\
\phi \sim N(0.5, 0.25) \\
\alpha_{pipe_i} \sim N(\bar{\alpha}, \sigma_\alpha) \\
\bar{\alpha} \sim T(4.2, 2.5, 3) \\
\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3) \\ 
\beta_j \sim T(0, 2.5, 3) \\
b_j \sim N(0, \sigma_b) \\
\sigma_b \sim  Half \text{-} T(0, 2.5, 3) \\ 
\end{align}
$$

where $T$ denotes the Student t-distribution with time-varying mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$. The mean is modeled as the sum of smooth functions of time $f_j(t)$. The full model (Zones 1 and 2) included a pipe-specific intercept $\alpha_{pipe_i}$ and centered smooth terms, whereas the Zone 1 model included non-centered series-specific smooths and a global intercept ($\bar{\alpha}$ in place of $\alpha_{pipe_i}$ in equation $(1)$). The matrices $Z_j$ and $X_j$ represent the penalized and unpenalized basis functions comprising each of the $f_j(t)$, and $b_j$ and $\beta_j$ represent the penalized and unpenalized GAM coefficients. $Gamma$ and $N$ denote the gamma and normal distributions.

On the log scale, the time-varying mean in the full model was estimated as the sum of a global multi-year trend, a set of local multi-year trends modifying the global trend to better fit the data from each location, and a second set of local multi-year trends capturing deviations of the individual time series from the global and location-level trends (Figure S3a). Since orthophosphate was increased on different dates in Zones 1 and 2, we expressed time as days before and after the respective increases. The time-varying mean in the Zone 1 model was estimated as the sum of a global multi-year trend, a seasonal trend, and a set of local multi-year trends capturing deviations of the individual time series from the global and seasonal trends (Figure S4). In both models, the multi-year trends were estimated using thin-plate regression splines, and the Zone 1 model's seasonal trend was estimated using a cyclic cubic regression spline [@mgcv].

The instantaneous rate of change in mean log lead concentration was estimated using finite differences, as described in a recent paper [@trueman_comparing_2022]. Briefly, we generated posterior predictions of the global or location-level multi-year trend along a regular time sequence spanning the range of the data. Then, we repeated this process after adding a small $\delta$ to each value in the sequence. The difference between posterior predictions evaluated at $t$ and $t + \delta$, divided by $\delta$, approximates the first derivative of the smooth term.

### Sentinel homes

Lead concentrations in point-of-use samples, $y_i$, were described using a multilevel model [@mcelreath_statistical_2016]. That is, the change in lead release accompanying the orthophosphate dose increase was estimated after accounting for the effects of sample location and profile litre. The model is specified in equation (2),

$$
(2)~~
\begin{align}
\text{likelihood:} \\
log(y_i) | censored = 0 \sim T(\mu_i, \sigma) \\
log(y_i) | censored = 1 \sim T\text{-}CDF(\mu_i, \sigma) \\
\\
\text{model for }\mu_i\text{:} \\
\mu_i = \alpha_{site_j} + \gamma_{sample_k} + \beta R \\
\\
\text{priors:} \\
\sigma \sim Half \text{-} N(0,1) \\
\nu \sim Gamma(2, 0.1) \\
\beta \sim N(-0.8,0.3) \\
\\
\alpha_{site_j} \sim N(\bar{\alpha}, \sigma_{\alpha}),~\text{for}~j~\text{in}~1..9 \\
\bar{\alpha} \sim N(0,1) \\
\sigma_{\alpha} \sim Half\text{-}Cauchy(0,1) \\
\\
\gamma_{sample_k} \sim N(0, \sigma_{\gamma}),~\text{for}~k~\text{in}~1..45 \\
\sigma_{\gamma} \sim Half\text{-}Cauchy(0,1) \\
\end{align}
$$

where $T$ again denotes the Student t-distribution with mean $\mu$, standard deviation $\sigma$, and degrees-of-freedom $\nu$; $censored$ is a binary variable indicating whether the sample concentration was observed or left-censored (i.e., a nondetect). The parameters $\alpha_{site_j}$ and $\gamma_{sample_k}$ are random intercepts describing each unique site/profile litre combination, $R$ is a binary variable indicating the sampling round (i.e., before/after the dose increase), and $\beta$ is the difference between rounds. $Half\text{-}Cauchy$, and $T \text{-} CDF$ represent the half-Cauchy distribution and the Student t cumulative distribution function (i.e., $P(X \leq x)$). $T \text{-} CDF$ quantifies the probability that $y_i$ is less than the censoring limit on the log scale. Nondetects, then, inform the model without the need to replace them with imputed values.

The priors on $\bar{\alpha}$, $\bar{\gamma}$, $\sigma_{\alpha}$ , and $\sigma_{\gamma}$ are weakly informative, meaning that they discourage unrealistic parameter estimates [@gelman_bayes_2009]. The prior on $\beta$---the difference between lead concentration at the two orthophosphate doses---was determined using posterior predictions from the generalized additive model of pipe loop data, as described in the results section.

# Results and discussion

## Quantifying the effect of an orthophosphate dose increase

### Sentinel pipe racks

Lead release from pipe racks was relatively constant at 1 mg PO~4~ L^-1^ (Figure 1c). At this dose, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ at all times (d[Pb]/dt ~ 0, Figure 1a). Pipe racks, then, appear to have been successfully stabilized at the initial orthophosphate concentration.

```{r figure-preds, fig.cap="**Figure 1. (a)** The global multi-year smooth term representing the change in lead concentration across all pipe sections, and the local modifiers representing deviations from the global trend to better fit data from each pipe rack. Red highlighting indicates the portion of the trend where a 95% credible interval on its slope does not include zero, and the shaded grey region represents a 95% credible interval on the time-varying mean. Sample collection dates are indicated by vertical ticks on the x-axis. **(b)** Orthophosphate in treated water, by zone. **(c)** Time series of total lead in effluent from lead pipes at three locations. Fitted values from the hierarchical GAM are superimposed on the time series in bold. Ticks at the top and bottom of the panels represent values outside the plotting limits."}
knitr::include_graphics(here::here("figures/figure-1.png"))
```

An increase to 2 mg PO~4~ L^-1^ was followed by a decreasing trend in lead concentration (Figure 1a, c). That is, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a). The higher dose, then, appeared to provide additional protection against lead release. Across both zones and all three pipe racks, doubling the orthophosphate dose decreased geometric mean lead concentrations within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).

An additional decrease in lead release was particular to Zone 2 and not accounted for by the global trend (Figure 1a). A possible explanation was a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S3b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020]. This lowered aluminum in treated water (Figure S5), and a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022]. Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles. And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead. Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999], but these fractions were not measured in treated water.

The decrease in the location-specific trend representing Zone 2 followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b). Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b). Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%). Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect. That is, the hierarchical nature of the model allows us to control for an effect common to all groups to better understand an effect that occurred in only one group.

### Sentinel homes

We used the estimated year-over-year decrease in geometric mean lead release from pipe racks (`r round(abs(pcent_decrease$y))`%) as a prior probability for orthophosphate's effect on lead concentrations in the sentinel homes' tap water (Figure 2a). The prior probability reflects our state of knowledge before learning from the point-of-use data; on the natural log scale, an approximation of the prior difference estimate is $N(\mu = -0.8, \sigma = 0.3)$, where $N$ is a Gaussian with mean $\mu$ and standard deviation $\sigma$.

```{r figure-pou, fig.cap="**Figure 2. (a)**  Density plots show the estimated percent change in lead at the point of use, comparing sample profiles collected at 1 mg PO~4~ L^-1^ and again, approximately 1 year later, at 2 mg PO~4~ L^-1^. Model predictions generated using a flat prior (i.e., no prior knowledge of the effect of orthophosphate) are compared against those generated using a prior informed by the GAM. (*N* denotes the normal distribution.) **(b)** Lead at the point of use, paired by site and profile litre. Left-censored values (i.e., nondetects) are represented by the horizontal/vertical ticks and the grey-shaded region at the bottom left of the plot. The red diagonal line represents the estimated difference between lead concentrations at the two doses, and the red-shaded region represents a 95% credible interval on that estimate (generated using an informative $N(-0.8, 0.3)$ prior)."}
knitr::include_graphics(here::here("figures/figure-2.png"))
```

Geometric mean lead release at the high orthophosphate dose (2 mg PO~4~ L^-1^) was `r d$d_50`% of that at the low dose (1 mg PO~4~ L^-1^), with a 95% credible interval of `r d$d_2.5`--`r d$d_97.5`% (Figure 2b). The choice of prior had little influence on the difference estimate: the corresponding estimate obtained by using an uninformative prior---assigning equal probability to all orthophosphate treatment effect sizes, whether physically plausible or not---was `r dnp$d_50`%, with a 95% credible interval of `r dnp$d_2.5`--`r dnp$d_97.5`%.

These estimates are somewhat smaller than the one based on pipe rack data. Differences in the models are a factor, but differences in materials also matter. That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon. Data from sentinel homes, though, also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous. Corrosion of lead solder, for instance, may be accelerated by orthophosphate [@nguyen_acceleration_2011]. To capture these effects, pipe racks could easily be modified to include copper and lead solder.

## Quantifying the effect of an orthophosphate dose decrease

A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b). We used the sentinel pipe racks to determine the orthophosphate dose response in this zone. That is, we estimated the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and the effect of a subsequent decrease to 1.5 mg PO~4~ L^-1^. But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects. Seasonality was more complex in Zone 2, perhaps due to the inverse seasonality in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]). And since the dose increases occurred in late November and July in Zones 1 and 2 respectively, controlling explicitly for seasonality in the full model---encompassing both zones---was less important.

```{r figure-3, fig.cap="**Figure 3.** **(a)** The global multi-year trend in lead release; red highlighting indicates the portion of the trend where the 95% credible interval on its slope does not include zero. **(b)** Orthophosphate in Zone 1 treated water. **(c)** The seasonal smooth term in the GAM. In **(a)** and **(c)**, shaded grey regions span 95% credible intervals on the trends, and ticks on the x-axes represent sample collection dates. **(d)** Time series of total lead in effluent from lead pipes at the two locations in Zone 1. Fitted values from the hierarchical GAM are superimposed on the time series in bold. Ticks at the top and bottom of the panels represent values outside the plotting limits. **(e)** Water temperature in pipe rack effluent; points represent medians and error bars span the range of measurements, by date. A cyclic cubic spline [@pedersen_hierarchical_2019] is superimposed in blue."}
knitr::include_graphics(here::here("figures/figure-3.png"))
```

As in the full model, mean (log) lead concentrations were relatively constant at 1 mg PO~4~ L^-1^: at this dose, a 95% credible interval on the slope of the global multi-year trend always included 0 µg Pb L^-1^ d^-1^ (Figure 3a). An increase to 2 mg PO~4~ L^-1^ was followed here as well by a decreasing trend in lead concentrations.

Even after accounting for the seasonal variation in lead release, though, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a) and a 95% credible interval on the slope of the global trend that did not include zero. The intermediate dose, then, appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses. Six months after the orthophosphate dose reduction, the increase in geometric mean lead release was estimated at `r round(pcent_increase$y[1])`%, with a 95% credible interval of `r round(pcent_increase$ymin[1])`--`r round(pcent_increase$ymax[1])`%.

This result has implications for passivation-maintenance orthophosphate dosing strategies---that is, initiating treatment at a high orthophosphate dose to promote lead phosphate scale formation and then decreasing the dose once scale evolution has slowed [@schock1998lead]. Although lead solubility is predicted to increase with a decrease in orthophosphate, the effect on particulate lead is unclear: an established lead-phosphate scale, for instance, may be no less durable after a decrease in the orthophosphate dose. But while passivation/maintenance dosing has the potential to conserve phosphorus, it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022]. Here, the dose response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.

# Conclusion

Point-of-use sampling is necessary to accurately quantify lead release into drinking water. But lead service line replacement, incomplete participation by residents in sampling programs, and changes to premises plumbing make it difficult to monitor the effectiveness of corrosion control over time this way. And while no simple apparatus can reliable quantify human exposure to lead, sentinel pipe racks offer an alternative to point-of-use sampling for non-regulatory monitoring. Especially when installed at multiple locations across a water distribution network, sentinel pipe racks can be used to understand how both anticipated and unexpected changes in water quality impact lead concentrations. We used them here to estimate the effect on lead release of changes in orthophosphate dose and coagulation process. By partitioning the variation in lead concentrations hierarchically---estimating global and location-level trends---we were better able to control for seasonality or other potential confounders before quantifying the effects of interest.

# Acknowledgements

This work was supported by Mitacs, through the Mitacs Accelerate Program (Reference # IT23352), and NSERC, through an Industrial Research Chair program (Grant # IRCPJ: 349838-16). We acknowledge Halifax Water's water quality department for managing the pipe rack and residential sampling programs.

# Supporting information

Graphical and tabular summaries of water quality, figures summarizing smooth terms not appearing in the body of the paper, instructions provided to volunteer residents, and a photo of a prototype pipe rack.

# References

### 2023-05-15 R/01-clean-data.R

@@ -1,0 +1,7 @@
+
+# --------------------- description ---------------------
+
+# Utility data were cleaned according to the principles outlined in "Tidy data" 
+# (https://doi.org/10.18637/jss.v059.i10). All location information was removed.
+
+

### 2023-05-15 R/02-model-gam.R

@@ -1,0 +1,76 @@
+
+# --------------------- setup ---------------------
+
+library("brms")
+library("bgamcar1")
+library("tidyverse")
+library("lubridate")
+
+options(mc.cores = parallel::detectCores())
+
+# -------------------- functions --------------------
+
+make_model_in <- function(x, ...) {
+  x %>% 
+    group_by(zone, rack, pipe_number) %>% 
+    mutate(
+      date_yday = yday(date) / if_else(leap_year(date), 366, 365),
+      d_x = date_numeric - lag(date_numeric),
+      d_x = replace_na(d_x, 0),
+      d_x = d_x / 27, # normalize to most common time difference
+      ...
+    ) %>% 
+    ungroup() 
+}
+
+# --------------------- read ---------------------
+
+data <- read_csv(here::here("data-clean/model-in-gam.csv")) %>% 
+  # add d_x variable
+  make_model_in(log_value = log(value))
+
+data_full <- filter(data, date <= "2022-05-16") # for Zone 1 and 2 model
+data_z1 <- filter(data, zone == 1) # for Zone 1 model
+
+# --------------------- inputs ---------------------
+
+priors <- c(
+  prior(student_t(3, 0, 2.5), b),
+  prior(normal(.5, .25), ar, lb = 0, ub = 1)
+)
+
+# --------------------- fit ---------------------
+
+model_gam <- fit_stan_model(
+  file = here::here("models/gam-brms"),
+  seed = 2124536,
+  bform = bf(
+    log_value ~ 
+      1 + (1 | series) +
+      s(date_numeric, k = 6) +
+      s(date_numeric, by = location, m = 1) +
+      s(date_numeric, by = series, m = 1) + 
+      ar(time = date_numeric, gr = series)
+  ),
+  bdata = data_full,
+  bpriors = priors,
+  control = list(adapt_delta = .9),
+  save_warmup = FALSE
+)
+
+model_gam_z1 <- fit_stan_model(
+  file = here::here("models/gam-brms-z1"),
+  seed = 2124536,
+  bform = bf(
+    log_value ~ 
+      s(date_numeric, k = 15) +
+      s(date_yday, bs = "cc") +
+      s(date_numeric, series, bs = "fs", m = 1) +
+      ar(time = date_numeric, gr = series)
+  ),
+  bdata = data_z1,
+  bpriors = priors,
+  knots = list(date_yday = c(0, 1)),
+  control = list(adapt_delta = .9),
+  save_warmup = FALSE
+)

### 2023-05-15 R/03-inform-prior-ml.R

@@ -1,0 +1,147 @@
+
+# inform the prior for the sentinel site model, 
+# and generate difference estimates for paper.Rmd
+
+# --------------------- setup ---------------------
+
+source(here::here("R/02-model-gam.R"))
+source(here::here("R/helpers-figures/analysis-ph.R"))
+library("ggdist")
+library("testthat")
+
+tol <- 30 # window in days in which to consider data for tests on difference estimates
+
+# --------------------- predict ---------------------
+
+###################################################################################
+# how much is lead predicted to decrease in the first year after the dose increase?
+
+newdata <- tibble(date_numeric = c(0, 365))
+
+# predict from global term of GAM:
+
+post <- posterior_smooths(model_gam, "s(date_numeric,k=6)", newdata = newdata)
+
+d <- post[,2] - post[,1]
+
+# for ml prior:
+
+avg <- round(mean(d), 1)
+stdev <- round(sd(d), 1)
+
+# (for paper.Rmd)
+
+pcent_decrease <- median_qi(d) %>% 
+  mutate(across(starts_with("y"), ~ 100 * (exp(.x) - 1)))
+
+test_that("percent decrease is reasonable", {
+  
+  before <- data_full %>%
+    filter(
+      date_numeric > -tol,
+      date_numeric < 0
+    ) %>%
+    with(mean(value))
+  
+  after <- data_full %>%
+    filter(abs(date_numeric - 365) < tol) %>%
+    with(mean(value))
+  
+  d <- 100 * (after - before) / before
+  
+  expect_true(d > pcent_decrease$ymin) 
+  expect_true(d < pcent_decrease$ymax)
+  
+})
+
+############################################################################
+# how much is lead predicted to decrease due to the coagulation pH increase?
+
+# for paper.Rmd:
+
+newdata2 <- tibble(
+  date_numeric = c(ph_increase, 0),
+  location = "z2.r1",
+  series = NA
+)
+
+post2 <- posterior_smooths(
+  model_gam, 
+  "s(date_numeric,by=location,m=1)", 
+  newdata = newdata2[, c("date_numeric", "location")]
+)
+
+d2 <- post2[,2] - post2[,1]
+
+pcent_decrease2 <- ggdist::median_qi(d2) %>% 
+  mutate(across(starts_with("y"), ~ 100 * (exp(.x) - 1)))
+
[…]

### 2023-05-15 R/04-model-ml.R

@@ -1,0 +1,58 @@
+
+# -------------------- setup --------------------
+
+source(here::here("R/03-inform-prior-ml.R"))
+library("glue")
+
+model_in <- read_csv(here::here("data-clean/model-input.csv"))
+
+# -------------------- inputs --------------------
+
+seed <- 421508
+
+bform <- bf(log(value) | cens(censored) ~ round + (1 | location) + (1 | location:litre))
+
+# -------------------- prior --------------------
+
+priors <- c(
+  set_prior(glue("normal({avg}, {stdev})"), "b"), # see inform-prior-ml.R
+  prior(normal(0, 1), Intercept),
+  prior(cauchy(0, 1), sd),
+  prior(normal(0, 1), sigma)
+)
+
+# -------------------- functions --------------------
+
+fit_brm <- function(path, ...) {
+  brm(
+    bform,
+    data = model_in,
+    family = "student",
+    cores = 4,
+    seed = seed,
+    iter = 4000,
+    file = here::here(path),
+    file_refit = "on_change",
+    ...
+  )
+}
+
+# -------------------- prior simulation --------------------
+
+model_prior <- fit_brm(
+  path = "models/prior",
+  sample_prior = "only",
+  prior = priors
+)
+
+# -------------------- models --------------------
+
+model <- fit_brm(
+  path = "models/model",
+  prior = priors
+)
+
+model_noprior <- fit_brm(
+  path = "models/model-noprior"
+)
+

### 2023-05-15 R/05-pp-check-gam.R

@@ -1,0 +1,118 @@
+
+source("R/02-model-gam.R")
+library("janitor")
+library("posterior")
+library("testthat")
+
+# -------------------- pp-check --------------------
+
+# full:
+
+pp_check(model_gam, "ecdf_overlay", ndraws = 100) + 
+  coord_cartesian(xlim = c(-5, 10))
+
+# zone 1:
+
+pp_check(model_gam_z1, "ecdf_overlay", ndraws = 100) + 
+  coord_cartesian(xlim = c(-5, 10))
+
+# -------------------- autocorrelation --------------------
+
+# full:
+
+resid <- bind_cols(data_full, residuals(model_gam, incl_autocor = FALSE)) %>% 
+  clean_names()
+
+resid %>% 
+  with(acf(estimate))
+
+# zone 1:
+
+resid_z1 <- bind_cols(data_z1, residuals(model_gam_z1, incl_autocor = FALSE)) %>% 
+  clean_names()
+
+resid_z1 %>% 
+  with(acf(estimate))
+
+# -------------------- plot residuals --------------------
+
+# full:
+
+resid %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(rowid, estimate)) +
+  geom_point()
+
+params <- c("df" = median(as_draws_df(model_gam, "nu")$nu))
+
+resid %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(sample = estimate)) +
+  stat_qq(distribution = brms::qstudent_t, dparams = params) +
+  stat_qq_line(distribution = brms::qstudent_t, dparams = params)
+
+resid %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(estimate)) +
+  geom_histogram()
+
+# zone 1:
+
+resid_z1 %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(rowid, estimate)) +
+  geom_point()
+
+params_z1 <- c("df" = median(as_draws_df(model_gam_z1, "nu")$nu))
+
+resid_z1 %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(sample = estimate)) +
+  stat_qq(distribution = brms::qstudent_t, dparams = params_z1) +
+  stat_qq_line(distribution = brms::qstudent_t, dparams = params_z1)
+
+resid_z1 %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(estimate)) +
+  geom_histogram()
+
+# -------------------- loo --------------------
+
[…]

### 2023-05-15 R/06-pp-check-ml.R

@@ -1,0 +1,101 @@
+
+source("R/04-model-ml.R")
+library("janitor")
+library("bayesplot")
+library("ggdist")
+library("posterior")
+library("testthat")
+
+# -------------------- prior simulation --------------------
+
+post_prior <- withr::with_seed(1538, {posterior_predict(model_prior, ndraws = 50)})
+
+post_prior_cens <- post_prior %>% 
+  apply(2, function(x) if_else(x < log(.5), log(.5), x))
+
+post_prior_cens %>% 
+  t() %>% 
+  data.frame() %>% 
+  pivot_longer(everything()) %>% 
+  ggplot(aes(value, group = name)) + 
+  geom_density() + 
+  coord_cartesian(xlim = c(-1, 10))
+
+# -------------------- prior sensitivity --------------------
+
+list(
+  "prior (sim)" = model_prior,
+  "posterior" = model,
+  "flat prior" = model_noprior
+) %>% 
+  map_dfr(
+    ~ as_draws_df(.x, "b_round2_after") %>% 
+      as_tibble(),
+    .id = "type"
+  ) %>% 
+  ggplot(aes(exp(b_round2_after), col = type, fill = type)) + 
+  stat_halfeye(slab_alpha = .5, position = position_dodge(width = .2)) + 
+  scale_y_continuous(expand = expansion(add = c(0, .2)))
+
+# -------------------- autocorrelation --------------------
+
+resid <- bind_cols(model_in, residuals(model)) %>% 
+  clean_names()
+
+resid %>% 
+  filter(censored != "left") %>% 
+  with(acf(estimate))
+
+# -------------------- plot residuals --------------------
+
+resid %>% 
+  filter(censored != "left") %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(rowid, estimate, col = location, shape = censored)) +
+  geom_point()
+
+resid %>% 
+  filter(censored != "left") %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(sample = estimate)) +
+  geom_qq() +
+  geom_qq_line()
+
+resid %>% 
+  filter(censored != "left") %>% 
+  rowid_to_column() %>% 
+  ggplot(aes(estimate)) +
+  geom_histogram()
+
+# -------------------- loo --------------------
+
+model_loo <- loo(model)
+
+# -------------------- pp check --------------------
+
+post <- withr::with_seed(158, {posterior_predict(model, ndraws = 500)})
[…]

### 2023-05-15 R/07-make-figures.R

@@ -1,0 +1,62 @@
+
+# --------------------- setup ---------------------
+
+source("R/04-model-ml.R")
+source("R/helpers-figures/figure-funs.R")
+source("R/helpers-figures/analysis-ph.R")
+library("patchwork")
+library("assertr")
+library("ggtext")
+library("scales")
+library("janitor")
+library("posterior")
+library("devtools", include.only = "session_info")
+
+# --------------------- theme ---------------------
+
+theme_set(
+  theme_bw() + 
+    theme(
+      plot.tag = element_text(face = "bold"),
+      legend.position = "bottom",
+      legend.margin = margin(),
+      legend.box.margin = margin(),
+      strip.text = ggtext::element_markdown()
+    )
+)
+
+palette <- wesanderson::wes_palette("Zissou1", 6, "continuous")
+pal_short <- palette[c(6,4,3,1)]
+alpha <- .3
+
+striplabs <- as_labeller(c(
+  "z1.r1" = "Zone 1 (Rack 1)", "z1.r2" = "Zone 1 (Rack 2)", 
+  "z2.r1" = "Zone 2 (Rack 1)", "global" = "Global"
+))
+
+# --------------------- read ---------------------
+
+wq <- read_csv(here::here("data-clean/data-wq-si.csv"))
+
+finished_ortho <- read_csv("data-clean/finished-water-orthophosphate-zones-1-2.csv")
+
+# --------------------- figures ---------------------
+
+# n.b., these have to be run in order
+
+# paper
+
+source("R/helpers-figures/figure-1.R")
+source("R/helpers-figures/figure-2.R")
+source("R/helpers-figures/figure-3.R")
+source("R/helpers-figures/figure-toc.R")
+
+# supplement
+
+source("R/helpers-figures/figure-s3.R")
+source("R/helpers-figures/figure-s4.R")
+source("R/helpers-figures/figure-s5.R")
+
+#------------------ session info ------------------
+
+writeLines(capture.output(devtools::session_info()), "session-info.txt")

### 2023-05-15 R/08-make-summary-table.R

@@ -1,0 +1,62 @@
+
+# make water quality summary tables for paper.Rmd:
+
+# --------------------- setup ---------------------
+
+library("tidyverse")
+library("assertr")
+
+wq <- read_csv("data-clean/data-wq.csv") %>% 
+  mutate(
+    # phase 1 == 1 ppm PO4, phase 2 == 2 ppm PO4:
+    param = case_when(
+      param == "orthophosphate" & date_numeric < 0 ~ "orthophosphate (phase 1)", 
+      param == "orthophosphate" & date_numeric >= 0 ~ "orthophosphate (phase 2)",
+      TRUE ~ param
+    )
+  )
+
+input <- read_csv("data-clean/summary-table-input.csv")
+
+params_together <- c("dissolved inorganic carbon", "ph", "free cl", 
+                     "organic carbon", "dissolved chloride", "turb")
+
+remove_these <- c("collected volume", "do %")
+
+# --------------------- functions ---------------------
+
+summarize_wq <- function(x, ...) {
+  x %>% 
+    group_by(param, unit, ...) %>% 
+    summarize(
+      med = median(value),
+      lq = quantile(value, 0.25),
+      uq = quantile(value, 0.75),
+      pcens = mean(censored != "none")
+    ) %>% 
+    ungroup() %>% 
+    verify(pcens < .25)
+}
+
+# --------------------- summary table for paper ---------------------
+
+summary <- wq %>% 
+  filter(param %in% params_together) %>%
+  summarize_wq() %>% 
+  left_join(input) %>% 
+  select(param = param_table, med, lq, uq, unit = unit_table)
+
+# --------------------- summary table for supplement (separate zones) ---------------------
+
+summary_si <- wq %>% 
+  summarize_wq(zone) %>% 
+  filter(!param %in% remove_these) %>% 
+  left_join(input) %>% 
+  select(param = param_table, zone, med, lq, uq, unit = unit_table)
+
+# --------------------- write ---------------------
+
+write_csv(summary, "data-clean/summary-table.csv")
+write_csv(summary_si, "data-clean/summary-table-si.csv")
+  
+  

### 2023-05-15 R/09-render-paper.R

@@ -1,0 +1,22 @@
+
+library("tidyverse")
+
+# paper:
+
+path <- "Rmarkdown/paper.Rmd"
+path_temp <- str_replace(path,"(.+/)(.+)", "\\1_\\2")
+
+read_file(path) %>% 
+  # move arabic numberal refs after punctuation:
+  str_replace_all("\\s+(\\[@.*?\\])([.,;:])?", "\\2\\1") %>% 
+  write_file(path_temp)
+
+rmarkdown::render(path_temp)
+
+file.rename(str_replace(path_temp, "\\.Rmd", "\\.docx"), str_replace(path, "\\.Rmd", "\\.docx"))
+
+unlink(path_temp)
+
+# SI:
+
+rmarkdown::render("Rmarkdown/paper-si.Rmd")

### 2023-05-15 R/helpers-figures/analysis-ph.R

@@ -1,0 +1,15 @@
+
+#-------- dose increases ----------
+
+p_inc_pw <- as.Date("2021-07-29") # Zone 2
+
+#-------- identify pH increase ----------
+
+pidat <- read_csv(here::here("data-clean/pi-process-data.csv"))
+
+ph_increase <- pidat %>% 
+  mutate(ph = stats::filter(ph, rep(1/3, 3), method = "convolution")) %>% 
+  filter(ph >= 6.3, date > "2021-01-01") %>% 
+  summarize(date = difftime(min(date), p_inc_pw)) %>% 
+  pull(date) %>% 
+  as.numeric()

### 2023-05-15 R/helpers-figures/figure-1.R

@@ -1,0 +1,176 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure 1 ---------------------
+
+# dates and labels:
+
+data_full <- mutate(data_full, zone = paste("Zone", zone))
+
+p_increase <- 0
+
+# range of data for each group:
+
+limits <- data_full %>% 
+  group_by(location) %>% 
+  summarize(min = min(date_numeric), max = max(date_numeric)) %>% 
+  ungroup()
+
+# generate predictions
+
+preds_global <- local_slope(
+  data_full, model_gam, "date_numeric", epsilon = .1, 
+  smooth = "s(date_numeric,k=6)", 
+  pts = 400
+) %>% 
+  summarize_slopes(width = .95)
+
+preds_loc <- local_slope(
+  data_full, model_gam, "date_numeric", epsilon = 1, 
+  smooth = "s(date_numeric,by=location,m=1)",
+  g_var = "location",
+  pts = 400,
+  # adding next line because brms::posterior_smooths() requires that all variables used in 
+  # constructing AR term also included in newdata arg... even if they are not used in the smooth:
+  add_vars = list(series = rep(unique(data_full$series)[1:3], each = 400))
+) %>% 
+  summarize_slopes(location, width = .95) %>% 
+  mutate(
+    zone = str_extract(location, "(?<=z)\\d"),
+    zone = paste("Zone", zone),
+    rack = str_extract(location, "(?<=r)\\d")
+  )
+
+preds <- data_full %>% 
+  bind_cols(fitted(model_gam, incl_autocor = FALSE)) %>% 
+  mutate(
+    Estimate = exp(Estimate),
+    pipe = factor(pipe_number)
+  )
+
+preds_series <- conditional_smooths(model_gam, smooths = "s(date_numeric,by=series,m=1)")[[1]] %>% 
+  mutate(
+    zone = str_extract(series, "(?<=z)\\d"),
+    pipe_number = str_extract(series, "(?<=p)\\d"),
+    location = str_extract(series, ".+(?=\\.p)")
+  ) %>% 
+  rename_all(~ str_remove_all(.x, "__$"))
+
+# generate plot:
+
+annot_1 <- tibble(
+  x = 0, 
+  y = -Inf,
+  lab = "Dose increase<br>(1&ndash;2 mg PO<sub>4</sub> L<sup>-1</sup>)",
[…]

### 2023-05-15 R/helpers-figures/figure-2.R

@@ -1,0 +1,124 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure 2 ---------------------
+
+# part 1:
+
+d_psim <- as_tibble(as_draws_df(model_prior, "b_round2_after")) # prior simulation
+d_p <- as_tibble(as_draws_df(model, "b_round2_after"))
+d_np <- as_tibble(as_draws_df(model_noprior, "b_round2_after"))
+
+ylabs <- c( "Flat", "N(-0.8, 0.3)", "N(-0.8, 0.3)\n(prior simulation)")
+
+p1 <- tibble(
+  a = d_np$b_round2_after,
+  b = d_p$b_round2_after, 
+  c = d_psim$b_round2_after
+) %>% 
+  pivot_longer(everything(), names_to = "model") %>% 
+  ggplot(aes(-100 * (1 - exp(value)), model, col = model, fill = model)) + 
+  ggdist::stat_halfeye(slab_alpha = .3) + 
+  geom_text(
+    data = tibble(
+      x = 85,
+      y = letters[1:3],
+      labels = ylabs
+    ),
+    aes(x, y, label = labels),
+    size = 2.5, hjust = 1,
+    inherit.aes = FALSE
+  ) +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  ) +
+  scale_colour_manual(values = c("black", palette[6], "black")) +
+  scale_fill_manual(values = c("black", palette[6], "black")) +
+  guides(colour = "none", fill = "none") +
+  # scale_y_discrete(expand  = expansion(add = c(0.3, 0))) +
+  labs(
+    x = "% change in [Pb]",
+    y = "Density"
+  )
+
+# part 2:
+
+d <- as_draws_df(model, "b_round2_after") %>% 
+  summarize_draws(~ quantile(.x, probs = c(0.025, .1, .5, .9, 0.975)))
+
+p2 <- model_in %>%
+  pivot_wider(
+    id_cols = c(location, litre, sample), 
+    names_from = round, 
+    values_from = c(value, censored)
+  ) %>% 
+  mutate(
+    litre = str_replace(litre, "Flushed", "5"),
+    litre = as.numeric(str_extract(litre, "\\d"))
+  ) %>% 
+  clean_names() %>% 
+  # ggplot(aes(value_1_before, value_2_after, col = litre)) +
+  ggplot(aes(value_1_before, value_2_after)) +
[…]

### 2023-05-15 R/helpers-figures/figure-3.R

@@ -1,0 +1,193 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure 3 ---------------------
+
+p_inc_z1 <- as.Date("2020-11-30")
+p_dec_z1 <- as.Date("2022-05-16")
+
+# generate predictions
+
+epreds <- data_z1 %>%
+  with(crossing(
+    series = unique(series),
+    date = seq(min(date), max(date), by = "1 week")
+  )) %>% 
+  mutate(
+    date_numeric = as.numeric(difftime(date, p_inc_z1, units = "days")), # date of P dose increase
+    zone = str_extract(series, "(?<=z)\\d"),
+    rack = paste("Rack", str_extract(series, "(?<=r)\\d")),
+    pipe_number = str_extract(series, "(?<=p)\\d")
+  ) %>% 
+  verify(min(date_numeric) == min(data_z1$date_numeric)) %>% 
+  # verify that max date_numeric is equal to max date_numeric in data_z1 that is divisible by 7:
+  verify(max(date_numeric) == 7 * floor(max(data_z1$date_numeric) / 7)) %>% 
+  make_model_in() %>% 
+  mutate(
+    epreds = fitted(model_gam_z1, newdata = ., incl_autocor = FALSE)[, "Estimate"],
+    epreds = exp(epreds)
+  )
+
+smooths <- conditional_smooths(model_gam_z1, prob = 0.95)
+
+preds_global <- local_slope(
+  data_z1, model_gam_z1, "date_numeric", epsilon = .1, 
+  smooth = "s(date_numeric,k=15)", 
+  pts = 400
+) %>% 
+  summarize_slopes(width = .95)
+
+# generate plot:
+
+fig_3a <- preds_global %>%
+  mutate(date = date_numeric + p_inc_z1) %>% 
+  ggplot(aes(date, smooth_estimate)) + 
+  scale_color_manual(values = palette[6], labels = label_parse()) +
+  geom_vline(xintercept = p_inc_z1, linetype = 3) +
+  geom_vline(xintercept = p_dec_z1, linetype = 3) +
+  geom_ribbon(aes(ymin = smooth_lower, ymax = smooth_upper), alpha = alpha) +
+  geom_line() +
+  geom_line(
+    data = function(x) {
+      x %>% 
+        mutate(
+          smooth_estimate = if_else(
[…]

### 2023-05-15 R/helpers-figures/figure-funs.R

@@ -1,0 +1,92 @@
+
+library("testthat")
+
+# --------------------- functions ---------------------
+
+summarize_slopes <- function(x, width = .95, ...) {
+  lwr <- (1 - width) / 2
+  upr <- 1 - (1 - width) / 2
+  x %>% 
+    group_by(date_numeric, ...) %>% 
+    summarize(
+      across(
+        c(slope, smooth),
+        .fns = list(
+          estimate = median,
+          lower = ~ quantile(.x, lwr),
+          upper = ~ quantile(.x, upr)
+        )
+      )
+    ) %>% 
+    ungroup()
+}
+
+test_that("summarize_slopes returns expected quantiles", {
+  x <- 1:100
+  out <- tibble(
+    date_numeric = 1,
+    slope = x,
+    smooth = x - 1,
+  ) %>% 
+    summarize_slopes()
+  expect_equal(out$slope_estimate, median(x))
+  expect_equal(out$slope_lower, quantile(x, .025))
+  expect_equal(out$slope_upper, quantile(x, .975))
+  expect_equal(out$smooth_estimate, median(x - 1))
+  expect_equal(out$smooth_lower, quantile(x - 1, .025))
+  expect_equal(out$smooth_upper, quantile(x - 1, .975))
+})
+
+add_label <- function(labels, ...) {
+  geom_richtext(
+    data = labels,
+    aes(x = x, y = y, label = lab),
+    size = 2.5,
+    label.r = unit(0, "cm"),
+    label.size = 0, 
+    alpha = .9,
+    inherit.aes = FALSE,
+    vjust = "inward",
+    ...
+  )
+}
+
+add_ribbon <- function(d, l, u, ...) {
+  summ_args <- enquos(...)
+  geom_ribbon(
+    data = . %>%
+      reframe(x = range(!!!summ_args)) %>%
+      mutate(
+        ymin = exp(as.numeric(d[, l])) * x,
+        ymax = exp(as.numeric(d[, u])) * x
+      ),
+    aes(x = x, ymin = ymin, ymax = ymax),
+    inherit.aes = FALSE,
+    fill = palette[6], col = NA, alpha = .3
+  )
+}
+
+add_segment <- function(d, ...) {
+  summ_args <- enquos(...)
+  geom_segment(
+    data = . %>% 
+      summarize(
+        x = min(!!!summ_args),
+        xend = max(!!!summ_args)
+      ) %>% 
[…]

### 2023-05-15 R/helpers-figures/figure-s3.R

@@ -1,0 +1,50 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure S3 ---------------------
+
+preds_series <- conditional_smooths(model_gam, smooths = "s(date_numeric,by=series,m=1)")[[1]] %>% 
+  mutate(
+    pipe_number = str_extract(series, "(?<=p)\\d"),
+    location = str_extract(series, ".+(?=\\.p)")
+  ) %>% 
+  rename_all(~ str_remove_all(.x, "__$"))
+
+p1 <- preds_series %>% 
+  left_join(limits, by = "location") %>% 
+  filter(date_numeric >= min, date_numeric <= max) %>% 
+  ggplot(aes(date_numeric, estimate, col = pipe_number)) +
+  facet_wrap(vars(location), labeller = striplabs, ncol = 3) +
+  geom_line() + 
+  geom_rug(
+    data = data,
+    sides = "b",
+    inherit.aes = FALSE,
+    aes(x = date_numeric)
+  ) +
+  scale_color_manual(values = pal_short) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(
+    x = "Time (days)",
+    y = "Partial effect",
+    col = "Pipe"
+  )
+
+ph_increase_date <- ph_increase + p_inc_pw
+
+annot_1$x <- ph_increase_date
+annot_1$y <- 5.6
+annot_1$lab <- "pH = 6.3"
+
+p2 <- pidat %>% 
+  filter(date > "2021-01-01") %>% 
+  ggplot(aes(date, ph)) + 
+  geom_line() + 
+  add_label(labels = annot_1, nudge_x = 55, col = palette[5]) +
+  geom_vline(xintercept = ph_increase_date, col = palette[5]) + 
+  labs(x = NULL, y = "Coagulation pH")
+
+figure_s3 <- wrap_plots(p1, p2, widths = c(3,1)) + 
+  plot_annotation(tag_levels = "a")
+
+ggsave("figures/figure-s3.png", figure_s3, dev = "png", dpi = 600, width = 6.5, height = 2.5)

### 2023-05-15 R/helpers-figures/figure-s4.R

@@ -1,0 +1,26 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure S4 ---------------------
+
+striplabs <- as_labeller(c("1" = "Zone 1 (Rack 1)", "2" = "Zone 1 (Rack 2)"))
+
+figure_s4 <- smooths[[3]] %>%
+  mutate(
+    date = date_numeric + p_inc_z1,
+    rack = as.numeric(str_extract(series, "(?<=r)\\d")),
+    pipe = str_extract(series, "(?<=p)\\d$")
+  ) %>%
+  ggplot(aes(date, estimate__, col = pipe)) +
+  facet_wrap(vars(rack), labeller = striplabs) +
+  scale_y_continuous(expand = expansion(add = .1)) +
+  scale_color_manual(values = palette[c(1,3,4,6)]) +
+  geom_line() +
+  add_rug(date) +
+  labs(
+    x = NULL,
+    y = "Partial effect",
+    col = "Pipe"
+  )
+
+ggsave("figures/figure-s4.png", figure_s4, dev = "png", dpi = 600, width = 6.5, height = 3)

### 2023-05-15 R/helpers-figures/figure-s5.R

@@ -1,0 +1,25 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure S5 ---------------------
+
+figure_s5 <- wq %>%
+  filter(param == "total aluminum", zone == 2, date < "2022-01-01") %>% 
+  mutate(
+    year = as.factor(year(date)),
+    yday = yday(date) / if_else(leap_year(date), 366, 365),
+    yday = 365 * yday + as.Date("2021-12-31")
+  ) %>% 
+  ggplot(aes(yday, value, col = year, group = interaction(series, year))) + 
+  scale_x_date(date_labels = "%b") +
+  scale_color_manual(values = palette[c(3,1,4,6)]) +
+  geom_point() +
+  geom_line() + 
+  ylim(c(0, 210)) +
+  labs(
+    x = NULL,
+    y = expression("[Al] (µg L"^-1*")"),
+    col = NULL
+  )
+
+ggsave("figures/figure-s5.png", figure_s5, dev = "png", dpi = 600, width = 6.5, height = 3)

### 2023-05-15 R/helpers-figures/figure-toc.R

@@ -1,0 +1,78 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- toc art ---------------------
+
+figure_toc <- preds_global %>%
+  mutate(
+    date = date_numeric + p_inc_z1,
+    smooth_estimate_nzd = if_else(
+      sign(slope_lower) == sign(slope_upper),
+      smooth_estimate,
+      NA_real_
+    ),
+    # create a grouping vector for the derivative label colours:
+    # start and end of regions of change:
+    start = !is.na(smooth_estimate_nzd) & is.na(lag(smooth_estimate_nzd)),
+    end = !is.na(smooth_estimate_nzd) & is.na(lead(smooth_estimate_nzd)),
+    smgrp = cumsum(start) + cumsum(end),
+    smgrp = if_else(smgrp %in% c(2, 4), 0, smgrp),
+    smgrp = fct_recode(
+      as.factor(smgrp),
+      "no trend" = "0",
+      "negative" = "1",
+      "positive" = "3"
+    )
+  ) %>% 
+  ggplot(aes(date, smooth_estimate)) + 
+  scale_color_manual(values = c("black", palette[c(1, 6)])) +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
+  ) +
+  geom_vline(xintercept = p_inc_z1, linetype = 3) +
+  geom_vline(xintercept = p_dec_z1, linetype = 3) +
+  geom_label(
+    data = . %>% 
+      distinct(smgrp) %>% 
+      mutate(
+        x = c(as.Date("2019-11-01"), as.Date("2021-05-01"), as.Date("2022-05-05")),
+        y = c(1.7, -0.7, -0.05),
+        label = c(
+          # two calls to atop() decreases spacing between lines:
+          "atop(NA, atop(textstyle(Orthophosphate),textstyle(dose)))", 
+          "frac(d*'['*Pb*']',dt)<0", "frac(d*'['*Pb*']',dt)>0"
+        )
+      ),
+    aes(x, y, label = label, col = smgrp),
+    label.r = unit(0, "lines"),
+    label.size = 0,
+    alpha = .8,
+    size = 2.5,
+    parse = TRUE,
+    show.legend = FALSE
+  ) +
+  geom_ribbon(aes(ymin = smooth_lower, ymax = smooth_upper), alpha = alpha) +
+  geom_line() +
+  geom_line(
[…]

### 2023-05-16 README.Rmd
update readme with preprint DOI

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

# cct-monitoring

This repository contains the code and data necessary to reproduce the main results of the preprint ["Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"](https://doi.org/10.26434/chemrxiv-2023-h1459).

### 2023-05-16 R/helpers-figures/figure-1.R
code cleanup

@@ -78,7 +78,6 @@ fig_1a <- preds_loc %>%
   bind_rows(mutate(preds_global, location = "global")) %>% 
   ggplot(aes(date_numeric, smooth_estimate)) +
   facet_wrap(vars(location), labeller = striplabs, ncol = 1) + 
-  # scale_color_manual(values = palette[6], labels = label_parse()) +
   scale_color_manual(values = palette[6]) +
   geom_vline(
     data = tibble(location = c("global", "z2.r1"), date_numeric = c(0, ph_increase)),
@@ -99,7 +98,6 @@ fig_1a <- preds_loc %>%
           )
         )
     },
-    # aes(col = "frac(d*'['*Pb*']',dt)<0")
     aes(col = "d[Pb]/dt < 0")
   ) +
   geom_rug(

### 2023-05-16 R/helpers-figures/figure-2.R

@@ -36,7 +36,6 @@ p1 <- tibble(
   scale_colour_manual(values = c("black", palette[6], "black")) +
   scale_fill_manual(values = c("black", palette[6], "black")) +
   guides(colour = "none", fill = "none") +
-  # scale_y_discrete(expand  = expansion(add = c(0.3, 0))) +
   labs(
     x = "% change in [Pb]",
     y = "Density"
@@ -58,7 +57,6 @@ p2 <- model_in %>%
     litre = as.numeric(str_extract(litre, "\\d"))
   ) %>% 
   clean_names() %>% 
-  # ggplot(aes(value_1_before, value_2_after, col = litre)) +
   ggplot(aes(value_1_before, value_2_after)) +
   annotate("label", 5, 30, label = "Litre 1", label.size = 0, alpha = .8, size = 2.5) +
   geom_point(
@@ -105,8 +103,6 @@ p2 <- model_in %>%
   ) +
   scale_x_log10() + 
   scale_y_log10() + 
-  # scale_color_gradient(high = "black", low = "grey", labels = c(1:4, "Flushed"), trans = "reverse") +
-  # scale_color_gradientn(colours = palette[-6], labels = c(1:4, "Flushed"), trans = "reverse") +
   theme(
     axis.title.x = ggtext::element_markdown(),
     axis.title.y = ggtext::element_markdown()

### 2023-05-16 R/helpers-figures/figure-3.R

@@ -58,7 +58,6 @@ fig_3a <- preds_global %>%
           )
         )
     },
-    # aes(col = "frac(d*'['*Pb*']',dt)!=0")
     aes(col = "d*'['*Pb*']'*'/'*dt!=0")
   ) +
   add_rug(date) +

### 2023-05-24 Rmarkdown/paper.Rmd
updates for journal submission

@@
 ---
-title: "Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"
+title: "Sentinel lead pipe racks quantify orthophosphate's dose-response"
 author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
@@
 
-Orthophosphate is used widely to control lead release from plumbing into tap water.
-Its effect can be difficult to quantify, though, since tap water lead concentrations are site-specific.
-Sentinel homes with lead service lines are ideal for evaluating orthophosphate corrosion control programs, but best practices dictate the removal of lead service lines once they are identified.
-Sentinel homes, then, often have too short a useful life to be used effectively.
-Here we explore an alternative: sentinel pipe racks constructed with recovered lead pipes and supplied with water directly from the distribution system.
-We also propose a strategy for analyzing pipe rack data based on the generalized additive model, which approximates time series as a sum of smooth functions.
-In this study, geometric mean lead release from pipe racks exhibited a pronounced dose response, falling by `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%) after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_increase, c(ymin, ymax)))))), collapse = "--")`%) after a decrease to 1.5 mg PO~4~ L^-1^.
-Data from the sentinel homes were largely consistent with those from pipe racks: geometric mean lead levels at the high orthophosphate dose (2 mg L^-1^) were `r d$d_50`% of those at the low dose (1 mg L^-1^, 95% credible interval: `r d$d_2.5`--`r d$d_97.5`%).
+Orthophosphate is used to minimize lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are plumbing-dependent.
+Homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate the removal of lead once identified, which complicates sampling plans.
+Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system.
+We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as a sum of smooth functions.
+In this study, geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^.
+Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose.
 Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring.
-They also provide a fully Bayesian framework for quantifying orthophosphate's effect on lead release that is well-suited to incorporating information from multiple sources.
+They also provide a Bayesian framework for quantifying orthophosphate's effect on lead release that can incorporate information from multiple sources.
 
 *Keywords:* EPA LCRR; Health Canada; corrosion control; generalized additive model; Bayesian multilevel model
+
+# Synopsis
+
+Sentinel pipe racks can be used to quantify the impact of planned and unplanned water quality changes on lead release.
+
+# Graphical abstract
 
@@
 
-It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics.
-And while lead solubility modeling can be informative, it fails to account for the complex mineralogy of lead corrosion scale or the generation of particles [@tully_water_2019; @trueman_characterizing_2019].
+It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022].
+And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with a few notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019].
 A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022].
@@
 The matrices $Z_j$ and $X_j$ represent the penalized and unpenalized basis functions comprising each of the $f_j(t)$, and $b_j$ and $\beta_j$ represent the penalized and unpenalized GAM coefficients.
+The parameter $\phi$ is the first-order autoregressive coefficient, and $s$ represents the spacing in time between consecutive observations.
 $Gamma$ and $N$ denote the gamma and normal distributions.

### 2023-05-25 Rmarkdown/paper.Rmd
added/edited refs

@@
 ---
-title: "Sentinel lead pipe racks quantify orthophosphate's dose-response"
+title: "Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water"
 author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
@@
 It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022].
-And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with a few notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019].
+And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019].
 A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022].
@@
 
-Here we describe an alternative: sentinel lead pipe racks operated with feedwater directly from the distribution system.
+Here we describe an alternative: sentinel lead pipe racks supplied with drinking water directly from the distribution system.
 While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible.
@@
 Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^.
-Our results provide a fully Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.
+Our results provide a Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.
 
@@
 The higher dose, then, appeared to provide additional protection against lead release.
-Across both zones and all three pipe racks, doubling the orthophosphate dose decreased geometric mean lead concentrations within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
+Across both zones and all three pipe racks, doubling the orthophosphate dose decreased the geometric mean lead concentration within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
 
@@
 And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead.
-Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999], but these fractions were not measured in treated water.
+Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999].
+These fractions, though, were not measured in treated water.
 
@@
 Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%).
-Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect.
+And since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect.
 That is, the hierarchical nature of the model allows us to control for an effect common to all groups to better understand an effect that occurred in only one group.
@@
 Differences in the models are a factor, but differences in materials also matter.
-That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon.
+That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon [@schock1996corr].
 Data from sentinel homes, though, also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous.

### 2023-05-26 README.Rmd
first journal submission

@@
 
-This repository contains the code and data necessary to reproduce the main results of the preprint ["Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"](https://doi.org/10.26434/chemrxiv-2023-h1459).
+This repository contains the code and data necessary to reproduce the main results of the preprint ["Sentinel lead pipe racks quantify orthophosphate’s dose-response in drinking water"](https://doi.org/10.26434/chemrxiv-2023-h1459).

### 2023-05-26 Rmarkdown/paper-si.Rmd
first journal submission

@@
 ---
-title: "*Supplementary information for:* Sentinel pipe racks quantify orthophosphate's effect on lead release into drinking water"
+title: "*Supplementary information for:* Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water"
 author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"

### 2023-05-26 Rmarkdown/paper.Rmd
first journal submission

@@
 We used R [@R], and a collection of contributed packages [@rmarkdown; @tidyverse; @readxl; @janitor; @assertr; @brms; @ggdist; @glue; @posterior; @testthat; @bayesplot; @patchwork; @ggtext; @devtools; @mgcv], to analyze and visualize the data.
-The code and data necessary to reproduce the main results of the paper are available online [@cct_monitoring].
+Materials (R code and data) necessary to reproduce the main results of the paper are available online [@cct_monitoring].
 
@@
 \text{likelihood:} \\
-log(y_i) | censored = 0 \sim T(\mu_i, \sigma) \\
-log(y_i) | censored = 1 \sim T\text{-}CDF(\mu_i, \sigma) \\
+log(y_i) | censored = 0 \sim T(\mu_i, \sigma, \nu) \\
+log(y_i) | censored = 1 \sim T\text{-}CDF(\mu_i, \sigma, \nu) \\
 \\
@@
 
-The priors on $\bar{\alpha}$, $\bar{\gamma}$, $\sigma_{\alpha}$ , and $\sigma_{\gamma}$ are weakly informative, meaning that they discourage unrealistic parameter estimates [@gelman_bayes_2009].
-The prior on $\beta$---the difference between lead concentration at the two orthophosphate doses---was determined using posterior predictions from the generalized additive model of pipe loop data, as described in the results section.
+The priors on $\bar{\alpha}$, $\sigma_{\alpha}$ , and $\sigma_{\gamma}$ are weakly informative, meaning that they discourage unrealistic parameter estimates [@gelman_bayes_2009].
+The prior on $\beta$---the difference between lead concentration at the two orthophosphate doses---was determined using posterior predictions from the generalized additive model of pipe loop data, as described in the Results and discussion.
 

### 2023-08-15 Rmarkdown/paper-si.Rmd
revisions post-review

@@
 ---
-title: "*Supplementary information for:* Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water"
-author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
+title: "*Supplementary information for:* Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control"
+author: "Benjamin F. Trueman^§,^\\*, Javier A. Locsin^§^, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
 date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
@@
 library("posterior")
-wq <- read_csv(here::here("data-clean/summary-table.csv"))
-wq_si <- read_csv(here::here("data-clean/summary-table-si.csv"))
+library("here")
+wq <- read_csv(here("data-clean/summary-table.csv"))
+wq_si <- read_csv(here("data-clean/summary-table-si.csv"))
 ```
@@
 
-This document has 5 pages, 5 figures, and 1 table.
+This document has X pages, X figures, and X tables.
 
@@
 wq_si %>% 
+  group_by(param) %>% 
+  mutate(unit = unfill_vec(unit)) %>% 
+  ungroup() %>% 
   transmute(
     Parameter = unfill_vec(param),
-    Unit = unfill_vec(str_replace(unit, "ug", "µg")),
+    Unit = str_replace(unit, "ug", "µg"),
     Zone = zone,
@@
 
-```{r figure-s2, fig.cap="**Figure S2.** An example instruction sheet distributed to volunteer residents collecting point-of-use samples from sentinel homes."}
+```{r figure-s2, fig.width=6.5, fig.height=5, fig.cap="**Figure S2.** X-ray diffraction patterns characterizing lead pipe corrosion scale recovered from sites 1--3 (a--c, respectively), by layer (L1--L3/4)."}
 knitr::include_graphics(here::here("figures/figure-s2.png"))
@@
 
-```{r figure-s3, fig.cap="**Figure S3.
-(a)** In the full (Zones 1 and 2) model, local multi-year smooths capturing the deviations of each series from the global and location-specific smooths.
-**(b)** Coagulation pH at the treatment plant supplying Zone 2."}
+```{r figure-s3, fig.cap="**Figure S3.** An example instruction sheet distributed to volunteer residents collecting point-of-use samples from sentinel homes."}
 knitr::include_graphics(here::here("figures/figure-s3.png"))
@@
 
-```{r figure-s4, fig.cap="**Figure S4.** In the Zone 1 model, local multi-year smooths capturing the deviations of each series from the global and seasonal smooths."}
+```{r figure-s4, fig.cap="**Figure S4.
+(a)** In the full (Zones 1 and 2) model, local multi-year smooths capturing the deviations of each series from the global and location-specific smooths.
+**(b)** Coagulation pH at the treatment plant supplying Zone 2."}
 knitr::include_graphics(here::here("figures/figure-s4.png"))
@@
 
-```{r figure-s5, fig.cap="**Figure S5.** Total aluminum in pipe rack effluent."}
+```{r figure-s5, fig.cap="**Figure S5.** In the Zone 1 model, local multi-year smooths capturing the deviations of each series from the global and seasonal smooths."}
 knitr::include_graphics(here::here("figures/figure-s5.png"))
 ```
+
+```{r figure-s6, fig.cap="**Figure S6.** Total aluminum in pipe rack effluent."}
+knitr::include_graphics(here::here("figures/figure-s6.png"))
+```
+
+# Lead pipe scale analysis
+
+This section describes our procedure for identifying crystalline phases in three lead pipe samples recovered from three sites in Zone 2.
+
+## Scale extraction
+
+Lead pipe samples were excavated carefully, sealed at either end, and delivered to the laboratory.
+We cut a 0.2 m long section from each sample with a hacksaw, and we cut each section in half longitudinally to reveal the interior.
+Pipe interiors were photographed, and corrosion scale was collected with a stainless steel spatula in discrete layers, differentiated by colour, texture, and distance from the pipe wall.
+Pipe samples yielded distinct outer (L1), middle (L2), and inner (L3--L4) layers.
+The collected scale was finely ground with a mortar and pestle, which was cleaned with dilute nitric acid and pure water between samples.
+
+## X-ray diffraction
+
+We identified crystalline phases in the scale samples using a Rigaku Ultima IV X-ray diffractometer with a copper K$\alpha$ radiation source.
+The X-ray tube was operated at 35 kV and 30 mA.
+Scans were acquired over the range 10 - 70&deg; (2$\theta$) with a step size of 0.04&deg; and a scan speed of 0.8 &deg; min^-1^.
+Diffraction patterns were compared against the standard powder diffraction files listed in Table S2.
+
+```{r table-s2}
+
+tibble::tribble(
+  ~Phase, ~`PDF entry no.`,
+  "Chloropyromorphite", "96-900-4202",
+  "Hydroxypyromorphite", "00-086-0236",
+  "Hydrocerussite", "96-201-2741",
+  "Cerussite", "00-047-1734",
+  "Litharge", "00-085-1739",
+  "Lead", "96-900-8478"
+) %>% 
+  knitr::kable(caption = "**Table S2.** Standard X-ray diffraction patterns and their powder diffraction file entry numbers.")
+
+```
+
+## Results
+
+Corrosion scale in the first pipe was highly crystalline (Figure 3a).
+The top two scale layers were dominated by a pyromorphite mineral, and the bottom layer contained hydrocerussite and pyromorphite.
+While the best-matching pyromorphite mineral was chloropyromorphite (Pb~3~(CO~3~)~2~(OH)~2~), standard diffraction patterns for chloro- and hydroxyl- pyromorphites are quite similar, and our match to chloro-pyromorphite is tentative.
+We identified it based on two key differences: 1) between 29.5 and 30.5 degrees (2$\theta$), chloropyromorphite yields two peaks to hydroxypyromorphite's one, and between 45 and 47 degrees, chloropyromorphite yields three clearly separated peaks to hydroxypyromorphite's two [@lytle_inhibition_2009].
+
+Pipes extracted at the other two sites 2--3 had similar corrosion scale to that of site 1 (Figure 3b--c, respectively).
+A pyromorphite mineral was present, although the match to chloropyromorphite was evenless certain here than in the XRD patterns representing site 1.
+XRD patterns also matched the lead carbonates hydrocerussite (sites 2--3) and cerussite (site 2), as well as the lead oxide litharge (site 3).
+
+# References

### 2023-08-15 Rmarkdown/paper.Rmd
revisions post-review

@@
 ---
-title: "Sentinel lead pipe racks quantify orthophosphate's dose-response in drinking water"
-author: "Benjamin F. Trueman^§,^\\*, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
+title: "Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control"
+author: "Benjamin F. Trueman^§,^\\*, Javier A. Locsin^§^, Wendy H. Krkošek^†^, and Graham A. Gagnon^§^"
 date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, Nova Scotia, Canada B3H 4R2  \n&nbsp;  \n^†^Halifax Water, 450 Cowie Hill Rd., Halifax, Nova Scotia, Canada,  B3P 2V3  \n&nbsp;   \n&nbsp;*Corresponding author  \n&nbsp;E-mail: benjamin.trueman@dal.ca    \n&nbsp;Tel:902.494.6070  \n&nbsp;Fax: 902.494.3105"
@@
 ---
+
+<!-- to do: -->
+<!-- 1. cite https://doi.org/10.1080/00031305.2018.1549100 for bayesian r2? -->
 
@@
 
-Orthophosphate is used to minimize lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are plumbing-dependent.
-Homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate the removal of lead once identified, which complicates sampling plans.
-Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system.
-We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as a sum of smooth functions.
+Orthophosphate can control lead release into tap water, but its benefits are difficult to quantify since lead concentrations are so plumbing-dependent.
+Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified.
+The best sentinel homes, then, are often short-lived.
+Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system.
+We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions.
 In this study, geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^.
@@
 Our results demonstrate sentinel pipe racks as a viable alternative to at-the-tap sampling for non-regulatory corrosion control monitoring.
-They also provide a Bayesian framework for quantifying orthophosphate's effect on lead release that can incorporate information from multiple sources.
+They also provide a Bayesian framework for quantifying changes in lead release that can incorporate information from multiple sources.
 
@@
 
-It can be difficult, though, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022].
-And while modeling can be informative, it generally fails to account for the complex mineralogy of lead corrosion scale or---with notable exceptions [@abokifa_modeling_2017]---the generation of particles [@tully_water_2019; @trueman_characterizing_2019].
+It can be difficult, however, to estimate orthophosphate's effect on lead in drinking water since lead concentrations are determined by site-specific plumbing characteristics [@redmon_lead_2022].
+Modeling can be informative, but it generally fails to account for the complex mineralogy of lead corrosion scale or---with notable exceptions [@abokifa_modeling_2017]---the release of particles [@mcneill_importance_2004; @tully_water_2019; @trueman_characterizing_2019].
 A decrease in tap water lead sampled at sentinel homes over time is the most reliable metric of orthophosphate's success, and homes supplied by lead service lines represent the population most at-risk [@healthcanada2022].
@@
 And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022].
+As highlighted previously [@masters_inherent_2016], they allow complete control of the flow rate, flow pattern, and stagnation time without involving or endangering consumers.
 
@@
 
-Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with a pooled median pH and dissolved inorganic carbon concentration in pipe rack effluent of 7.3 and 4.4 mg C L^-1^ (Table 1).
-And while water quality in Zones 1 and 2 was largely similar, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022].
+Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with median pH and dissolved inorganic carbon concentrations in pipe rack effluent of 7.3--7.4 and 4--5 mg C L^-1^ (Table 1).
+But while water quality in Zones 1 and 2 was largely similar, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022].
 Aluminum concentrations in Zone 1 were much lower and more consistent throughout the year (Table S1).
@@
 wq %>% 
+  group_by(param) %>% 
+  mutate(unit = unfill_vec(unit)) %>% 
+  ungroup() %>% 
   transmute(
-    Parameter = param,
+    Parameter = unfill_vec(param),
     Unit = str_replace(unit, "ug", "µg"),
+    Zone = zone,
     Median = round(med, 1), 
@@
   ) %>% 
-  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent; these pooled estimates represent both zones (zone-specific water quality is summarized in Table S1).")
+  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent by zone, with the exception of hardness, which was measured in distributed water samples.")
 ```
@@
 Pipe racks were installed in utility-owned infrastructure; two were located in Zone 1 and one in Zone 2.
-Each was fitted with four replicate recovered lead pipe sections, supplied in parallel with water from the distribution system (an example is shown in Figure S1).
-Each pipe was excavated and handled according to principles outlined in a recent paper [@harmon_holistic_2022] and was approximately 60 cm long with an internal diameter of 1.3 cm.
-Each was connected to plastic tubing at either end with a brass compression fitting, yielding two galvanic lead-brass connections per pipe.
+In Zone 1, one site was below-ground and one above-ground, while the Zone 2 site was below-ground.
+Each rack was fitted with four replicate recovered lead pipe sections, supplied in parallel with water from the distribution system (Figure S1).
+Pipes were excavated and handled according to principles outlined in a recent paper [@harmon_holistic_2022] and were approximately 60 cm long with internal diameters of 1.3 cm.
+Pipe sections were connected to plastic tubing at either end with a brass compression fitting, yielding two galvanic lead-brass connections each.
 A timed valve supplied flow to the pipe sections for two minutes every six hours, and samples were collected approximately monthly, as the valves opened, at a nominal flow rate of 8 L min^-1^.
+Effluent not captured for analysis was discharged to a drain.
+Throughout, we used wide-mouth, plastic bottles and acidified samples with trace metal-grade nitric acid.
+
+Since the pipe racks remain in use at the time of writing, scale analysis of the recovered pipe sections was not possible.
+Corrosion scale representing lead pipes recovered from Zone 2 using the same method was characterized by 3--4 layers, featuring a mixture of pyromorphite (Pb~5~(PO~4~)~3~-(Cl,OH)), hydrocerussite (Pb~3~(CO~3~)~2~(OH)~2~), cerussite (PbCO~3~), and litharge ($\alpha$-PbO) (Figure S2 and Table S2).
 
@@
 This 4 $\times$ 1L profile was followed first by a 10-minute flush of the plumbing and then by collection of a final 1L sample.
+All samples were collected in wide-mouth plastic bottles and acidified with trace metal-grade nitric acid.
+Volunteers were instructed to open their faucets all the way and not to adjust the flow rate while sampling (Figure S3).
+
 Sample profiles were collected in May--June 2021, at 1 mg PO~4~ L^-1^, and again in May--June 2022, at 2 mg PO~4~ L^-1^.
-An example instruction sheet provided to residents is included as Figure S2.
+An example instruction sheet provided to residents is included in the supplementary information (Figure S3).
 During the study, all residents were provided with pitcher filters certified by NSF for removal of lead.
@@
 An accredited laboratory measured lead, iron, manganese, zinc, and aluminum [@epa2014method], as well as dissolved inorganic and total organic carbon [@sm_5310], chloride [@sm_4500_Cl], sulfate [@astm_516], orthophosphate [@sm_4500_P], and alkalinity [@epa1974method] in pipe rack effluent samples.
-Turbidity, pH, free chlorine, temperature, conductivity, dissolved oxygen, and oxygen reduction potential were determined onsite using portable Hach instruments.
+Turbidity, pH, free chlorine, temperature, conductivity, dissolved oxygen, and oxygen reduction potential were determined onsite using portable instruments (Hach, Loveland, CO).
 Orthophosphate was also quantified [@sm_4500_P] in treated water by Zone 1 and 2 treatment plant staff.
@@
 
-On the log scale, the time-varying mean in the full model was estimated as the sum of a global multi-year trend, a set of local multi-year trends modifying the global trend to better fit the data from each location, and a second set of local multi-year trends capturing deviations of the individual time series from the global and location-level trends (Figure S3a).
+On the log scale, the time-varying mean in the full model was estimated as the sum of a global multi-year trend, a set of local multi-year trends modifying the global trend to better fit the data from each location, and a second set of local multi-year trends capturing deviations of the individual time series from the global and location-level trends (Figure S4a).
 Since orthophosphate was increased on different dates in Zones 1 and 2, we expressed time as days before and after the respective increases.
-The time-varying mean in the Zone 1 model was estimated as the sum of a global multi-year trend, a seasonal trend, and a set of local multi-year trends capturing deviations of the individual time series from the global and seasonal trends (Figure S4).
+The time-varying mean in the Zone 1 model was estimated as the sum of a global multi-year trend, a seasonal trend, and a set of local multi-year trends capturing deviations of the individual time series from the global and seasonal trends (Figure S5).
 In both models, the multi-year trends were estimated using thin-plate regression splines, and the Zone 1 model's seasonal trend was estimated using a cyclic cubic regression spline [@mgcv].
@@
 
-Lead concentrations in point-of-use samples, $y_i$, were described using a multilevel model [@mcelreath_statistical_2016].
-That is, the change in lead release accompanying the orthophosphate dose increase was estimated after accounting for the effects of sample location and profile litre.
+Lead concentrations in point-of-use samples, $y_i$, were described using a multilevel model [@mcelreath_statistical_2016], where the change in lead release accompanying the orthophosphate dose increase was estimated after accounting for the effects of sample location and profile litre.
 The model is specified in equation (2),
@@
 \text{likelihood:} \\
-log(y_i) | censored = 0 \sim T(\mu_i, \sigma, \nu) \\
-log(y_i) | censored = 1 \sim T\text{-}CDF(\mu_i, \sigma, \nu) \\
+log(y_i) | censored_i = 0 \sim T(\mu_i, \sigma, \nu) \\
+log(y_i) | censored_i = 1 \sim T\text{-}CDF(\mu_i, \sigma, \nu) \\
 \\
 \text{model for }\mu_i\text{:} \\
-\mu_i = \alpha_{site_j} + \gamma_{sample_k} + \beta R \\
+\mu_i = \alpha_{site_j} + \gamma_{sample_k} + \beta R_i \\
 \\
 \text{priors:} \\
-\sigma \sim Half \text{-} N(0,1) \\
+\sigma \sim Half \text{-} N(0,0.5) \\
 \nu \sim Gamma(2, 0.1) \\
@@
 
-where $T$ again denotes the Student t-distribution with mean $\mu$, standard deviation $\sigma$, and degrees-of-freedom $\nu$; $censored$ is a binary variable indicating whether the sample concentration was observed or left-censored (i.e., a nondetect).
-The parameters $\alpha_{site_j}$ and $\gamma_{sample_k}$ are random intercepts describing each unique site/profile litre combination, $R$ is a binary variable indicating the sampling round (i.e., before/after the dose increase), and $\beta$ is the difference between rounds.
+where $T$ again denotes the Student t-distribution with mean $\mu_i$, standard deviation $\sigma$, and degrees-of-freedom $\nu$; $censored_i$ is a binary variable indicating whether the sample concentration was observed or left-censored (i.e., a nondetect).
+The parameters $\alpha_{site_j}$ and $\gamma_{sample_k}$ are random intercepts describing each unique site/profile litre combination, $R_i$ is a binary variable indicating the sampling round (i.e., before/after the dose increase), and $\beta$ is the difference between rounds.
 $Half\text{-}Cauchy$, and $T \text{-} CDF$ represent the half-Cauchy distribution and the Student t cumulative distribution function (i.e., $P(X \leq x)$).
-$T \text{-} CDF$ quantifies the probability that $y_i$ is less than the censoring limit on the log scale.
-Nondetects, then, inform the model without the need to replace them with imputed values.
+$T \text{-} CDF$ quantifies the probability that $y_i$ is less than the censoring limit on the log scale, allowing nondetects to inform the model without the need to replace them with imputed values.
 
@@
 
-Lead release from pipe racks was relatively constant at 1 mg PO~4~ L^-1^ (Figure 1c).
-At this dose, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ at all times (d[Pb]/dt ~ 0, Figure 1a).
-Pipe racks, then, appear to have been successfully stabilized at the initial orthophosphate concentration.
+Doubling the orthophosphate dose decreased lead release from pipe racks substantially.
+At 1 mg PO~4~ L^-1^, lead in pipe rack effluent did not exhibit a clear time-trend  (Figure 1c).
+In particular, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ everywhere (d[Pb]/dt ~ 0, Figure 1a).
+Pipe racks, then, appear to have stabilized well at the initial orthophosphate concentration.
 
@@
 **(c)** Time series of total lead in effluent from lead pipes at three locations.
-Fitted values from the hierarchical GAM are superimposed on the time series in bold.
+Fitted values from the hierarchical GAM are superimposed on the time series as bold lines.
 Ticks at the top and bottom of the panels represent values outside the plotting limits."}
@@
 
-An increase to 2 mg PO~4~ L^-1^ was followed by a decreasing trend in lead concentration (Figure 1a, c).
-That is, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a).
-The higher dose, then, appeared to provide additional protection against lead release.
-Across both zones and all three pipe racks, doubling the orthophosphate dose decreased the geometric mean lead concentration within a year by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
+A pronounced decreasing trend in lead concentration followed the increase to 2 mg PO~4~ L^-1^ (Figure 1a, c).
+Specifically, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a).
+The higher dose appeared to protect against lead release; and while this effect was slow to manifest, doubling the orthophosphate dose decreased the geometric mean lead concentration, year-over-year, by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
 
-An additional decrease in lead release was particular to Zone 2 and not accounted for by the global trend (Figure 1a).
-A possible explanation was a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S3b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020].
-This lowered aluminum in treated water (Figure S5), and a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022].
-Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles.
+Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding location-specific smooth term (Figure 1a).
+One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020].
+While this did not impact distributed water pH, it did lower aluminum in treated water (Figure S6).
+And a decrease in the aluminum concentration predicts a decrease in lead solubility---assuming that some fraction of dissolved aluminum precipitates with orthophosphate, leaving less available to react with lead [@trueman_seasonal_2022].
+Less aluminum in solution may also mean less post-precipitation of aluminum as particles and less adsorption of lead to those particles [@kvech_role_2001].
 And since suspended colloids containing aluminum and lead have been identified in Zone 2 [@trueman_seasonal_2022], the increase in coagulation pH may have decreased the capacity of distributed water to transport lead.
-Moreover, an improved coagulation process would be expected to remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999].
-These fractions, though, were not measured in treated water.
+Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999].
+But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data are not rich enough to identify the true cause.
 
-The decrease in the location-specific trend representing Zone 2 followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b).
+The decrease in the location-specific trend representing Zone 2, however, followed closely the increase in coagulation pH, and neither of the Zone 1 trends decreased comparably (Figure 1b).
 Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b).
 Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%).
-And since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the coagulation pH effect.
-That is, the hierarchical nature of the model allows us to control for an effect common to all groups to better understand an effect that occurred in only one group.
+Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the effect we attributed to a change in coagulation pH.
+That is, the hierarchical nature of the model allows us to control statistically for an effect common to all groups to better estimate an effect that occurred in only one group [@pedersen_hierarchical_2019].
 
@@
 Differences in the models are a factor, but differences in materials also matter.
-That is, pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon [@schock1996corr].
-Data from sentinel homes, though, also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous.
+Pipe racks measure the response of lead pipe to orthophosphate treatment, which tends to be quite large at slightly basic pH and low dissolved inorganic carbon [@schock1996corr].
+Data from sentinel homes also capture the effect of orthophosphate on lead release from other sources, which is much more ambiguous.
 Corrosion of lead solder, for instance, may be accelerated by orthophosphate [@nguyen_acceleration_2011].
 To capture these effects, pipe racks could easily be modified to include copper and lead solder.
+
+Even without an informative prior on $\beta$ (equation 2), the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks.
+This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016].
+And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we should expect lower variability [@masters_inherent_2016].
 
@@
 We used the sentinel pipe racks to determine the orthophosphate dose response in this zone.
-That is, we estimated the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and the effect of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
-But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects.
+We estimated the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and the effect of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
+But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016].
 Seasonality was more complex in Zone 2, perhaps due to the inverse seasonality in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]).
 And since the dose increases occurred in late November and July in Zones 1 and 2 respectively, controlling explicitly for seasonality in the full model---encompassing both zones---was less important.
+
+Temperature variation in the pipe rack effluent was a reasonable match for distributed water.
+While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked those of distributed water and point-of-use flushed samples more closely than the above-ground site (Rack 2), both captured the seasonal pattern well.
 
@@
 Ticks at the top and bottom of the panels represent values outside the plotting limits.
-**(e)** Water temperature in pipe rack effluent; points represent medians and error bars span the range of measurements, by date.
-A cyclic cubic spline [@pedersen_hierarchical_2019] is superimposed in blue."}
+**(e)** Water temperature in pipe rack effluent, by rack.
+Distributed water (water main) data and point-of-use flushed sample data [@trueman_evaluating_2016] are superimposed in colour, and trends are summarized by the solid and dashed lines as cyclic cubic splines [@pedersen_hierarchical_2019]."}
 knitr::include_graphics(here::here("figures/figure-3.png"))
@@
 
-Even after accounting for the seasonal variation in lead release, though, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a) and a 95% credible interval on the slope of the global trend that did not include zero.
-The intermediate dose, then, appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses.
+Even after accounting for the seasonal variation in lead release, though, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a) and a 95% credible interval on the slope of the global trend that excluded zero.
+The intermediate dose appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses.
 Six months after the orthophosphate dose reduction, the increase in geometric mean lead release was estimated at `r round(pcent_increase$y[1])`%, with a 95% credible interval of `r round(pcent_increase$ymin[1])`--`r round(pcent_increase$ymax[1])`%.
@@
 Although lead solubility is predicted to increase with a decrease in orthophosphate, the effect on particulate lead is unclear: an established lead-phosphate scale, for instance, may be no less durable after a decrease in the orthophosphate dose.
-But while passivation/maintenance dosing has the potential to conserve phosphorus, it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022].
-Here, the dose response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.
+Passivation/maintenance dosing has the potential to conserve phosphorus, but it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022].
+Here though, the dose response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.
 
@@
 
-Graphical and tabular summaries of water quality, figures summarizing smooth terms not appearing in the body of the paper, instructions provided to volunteer residents, and a photo of a prototype pipe rack.
+Graphical and tabular summaries of water quality, figures summarizing smooth terms not appearing in the body of the paper, instructions provided to volunteer residents, methods used to identify crystalline phases in samples of lead pipe scale (along with results), and a photo of a prototype pipe rack.
 

### 2023-08-15 R/06-pp-check-ml.R

@@ -85,7 +85,7 @@ ppc_stat(1 * (model_in$censored == "left"), 1 * (post < log(.5)), binwidth = .02
 
 # -------------------- diagnostic stats --------------------
 
-draws <- as_draws_df(model_gam) %>% 
+draws <- as_draws_df(model) %>% 
   summarize_draws()
 
 test_that("rhats are all less than 1.05", {

### 2023-08-15 R/07-make-figures.R

@@ -38,7 +38,17 @@ striplabs <- as_labeller(c(
 
 wq <- read_csv(here::here("data-clean/data-wq-si.csv"))
 
-finished_ortho <- read_csv("data-clean/finished-water-orthophosphate-zones-1-2.csv")
+finished_ortho <- read_csv(here::here("data-clean/finished-water-orthophosphate-zones-1-2.csv"))
+
+temp_z1_r2 <- read_csv(here::here("data-clean/temperature-ds-z1-r2.csv"))
+
+temp_flushed <- read_csv(here::here("data-clean/flushed-sample-temp.csv"))
+
+# for XRD data:
+
+xrd <- read_csv(here::here("data-clean/xrd-cleaned.csv"))
+
+stds <- read_csv(here::here("data-clean/xrd-standards.csv")) 
 
 # --------------------- figures ---------------------
 
@@ -53,9 +63,10 @@ source("R/helpers-figures/figure-toc.R")
 
 # supplement
 
-source("R/helpers-figures/figure-s3.R")
+source("R/helpers-figures/figure-s2.R")
 source("R/helpers-figures/figure-s4.R")
 source("R/helpers-figures/figure-s5.R")
+source("R/helpers-figures/figure-s6.R")
 
 #------------------ session info ------------------
 

### 2023-08-15 R/helpers-figures/figure-funs.R

@@ -90,3 +90,78 @@ add_rug <- function(...) {
     inherit.aes = FALSE
   )
 }
+
+# plot XRD data:
+
+gg_xrd <- function(pipe, layers = 3, phases, data, standards) {
+  
+  panels <- layers + length(phases)
+  
+  layer_names <- paste0("L", seq_len(layers)) %>% 
+    # pad layer names with NAs to reach length of 4
+    c(rep(NA_character_, 4 - layers))
+  
+  phase_names <- c(phases, rep(NA_character_, 6 - length(phases)))
+  
+  xrd_list <- list(
+    filter(data, code == pipe, layer == layer_names[1]),
+    filter(data, code == pipe, layer == layer_names[2]),
+    filter(data, code == pipe, layer == layer_names[3]),
+    filter(data, code == pipe, layer == layer_names[4]),
+    filter(standards, phase == phases[1]),
+    filter(standards, phase == phases[2]),
+    filter(standards, phase == phases[3]),
+    filter(standards, phase == phases[4]),
+    filter(standards, phase == phases[5]),
+    filter(standards, phase == phases[6])
+  )
+  
+  xrd_list <- xrd_list[map_lgl(xrd_list, ~ nrow(.x) > 0)]
+  
+  xrd_list %>% 
+    set_names(letters[1:panels]) %>% 
+    bind_rows(.id = "facet") %>% 
+    ggplot(aes(two_theta, intensity)) + 
+    facet_wrap(~ facet, ncol = 1, scales = "free_y") + 
+    geom_line(data = function(data) data %>% filter(is.na(phase)), linewidth = .2) + 
+    geom_segment(
+      data = function(data) data %>% filter(!is.na(phase)),
+      aes(x = two_theta, xend = two_theta, y = 0, yend = intensity), 
+      col = "deepskyblue4", linewidth = .25
+    ) +
+    geom_text(
+      data = function(data) data %>% 
+        filter(!is.na(phase)) %>% 
+        distinct(phase, facet),
+      aes(x = 55, y = .75, label = phase), 
+      col = "deepskyblue4", size = 3
+    ) +
+    geom_text(
+      data = function(data) data %>% 
+        filter(is.na(phase)) %>% 
+        group_by(phase, facet, layer) %>% 
+        summarize(y = 1.1 * intensity[which.max(two_theta)]),
+      aes(x = 73, y = y, label = layer), size = 2.5
+    ) +
+    labs(
[…]

### 2023-08-15 R/helpers-figures/figure-s2.R

@@ -1,0 +1,28 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure S2 ---------------------
+
+figure_s2 <- pmap(
+  list(
+    c("H1", "H2", "H3"), # site codes
+    c(3, 4, 3), # panels
+    c("Chloropyromorphite,Hydrocerussite,Cerussite,Lead",
+      "Chloropyromorphite,Hydrocerussite,Cerussite,Lead", 
+      "Chloropyromorphite,Hydrocerussite,Litharge,Lead")
+  ),
+  .f = function(x, y, z) {
+    gg_xrd(
+      pipe = x, 
+      layers = y,
+      phases = str_split(z, ",")[[1]],
+      data = xrd, 
+      standards = stds
+    )
+  }
+) %>% 
+  wrap_plots() + 
+  plot_annotation(tag_levels = "a") &
+  theme(plot.tag = element_text(face = "bold"))
+
+ggsave("figures/figure-s2.png", figure_s2, dev = "png", dpi = 600, width = 6.5, height = 6)

### 2023-08-15 R/helpers-figures/figure-s3.R

@@ -1,50 +1,0 @@
-
-# run this script from "07-make-figures.R"
-
-# --------------------- figure S3 ---------------------
-
-preds_series <- conditional_smooths(model_gam, smooths = "s(date_numeric,by=series,m=1)")[[1]] %>% 
-  mutate(
-    pipe_number = str_extract(series, "(?<=p)\\d"),
-    location = str_extract(series, ".+(?=\\.p)")
-  ) %>% 
-  rename_all(~ str_remove_all(.x, "__$"))
-
-p1 <- preds_series %>% 
-  left_join(limits, by = "location") %>% 
-  filter(date_numeric >= min, date_numeric <= max) %>% 
-  ggplot(aes(date_numeric, estimate, col = pipe_number)) +
-  facet_wrap(vars(location), labeller = striplabs, ncol = 3) +
-  geom_line() + 
-  geom_rug(
-    data = data,
-    sides = "b",
-    inherit.aes = FALSE,
-    aes(x = date_numeric)
-  ) +
-  scale_color_manual(values = pal_short) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-  labs(
-    x = "Time (days)",
-    y = "Partial effect",
-    col = "Pipe"
-  )
-
-ph_increase_date <- ph_increase + p_inc_pw
-
-annot_1$x <- ph_increase_date
-annot_1$y <- 5.6
-annot_1$lab <- "pH = 6.3"
-
-p2 <- pidat %>% 
-  filter(date > "2021-01-01") %>% 
-  ggplot(aes(date, ph)) + 
-  geom_line() + 
-  add_label(labels = annot_1, nudge_x = 55, col = palette[5]) +
-  geom_vline(xintercept = ph_increase_date, col = palette[5]) + 
-  labs(x = NULL, y = "Coagulation pH")
-
-figure_s3 <- wrap_plots(p1, p2, widths = c(3,1)) + 
-  plot_annotation(tag_levels = "a")
-
-ggsave("figures/figure-s3.png", figure_s3, dev = "png", dpi = 600, width = 6.5, height = 2.5)

### 2023-08-15 R/helpers-figures/figure-s5.R

@@ -3,23 +3,24 @@
 
 # --------------------- figure S5 ---------------------
 
-figure_s5 <- wq %>%
-  filter(param == "total aluminum", zone == 2, date < "2022-01-01") %>% 
+striplabs <- as_labeller(c("1" = "Zone 1 (Rack 1)", "2" = "Zone 1 (Rack 2)"))
+
+figure_s5 <- smooths[[3]] %>%
   mutate(
-    year = as.factor(year(date)),
-    yday = yday(date) / if_else(leap_year(date), 366, 365),
-    yday = 365 * yday + as.Date("2021-12-31")
-  ) %>% 
-  ggplot(aes(yday, value, col = year, group = interaction(series, year))) + 
-  scale_x_date(date_labels = "%b") +
-  scale_color_manual(values = palette[c(3,1,4,6)]) +
-  geom_point() +
-  geom_line() + 
-  ylim(c(0, 210)) +
+    date = date_numeric + p_inc_z1,
+    rack = as.numeric(str_extract(series, "(?<=r)\\d")),
+    pipe = str_extract(series, "(?<=p)\\d$")
+  ) %>%
+  ggplot(aes(date, estimate__, col = pipe)) +
+  facet_wrap(vars(rack), labeller = striplabs) +
+  scale_y_continuous(expand = expansion(add = .1)) +
+  scale_color_manual(values = palette[c(1,3,4,6)]) +
+  geom_line() +
+  add_rug(date) +
   labs(
     x = NULL,
-    y = expression("[Al] (µg L"^-1*")"),
-    col = NULL
+    y = "Partial effect",
+    col = "Pipe"
   )
 
 ggsave("figures/figure-s5.png", figure_s5, dev = "png", dpi = 600, width = 6.5, height = 3)

### 2023-08-15 R/helpers-figures/figure-s6.R

@@ -1,0 +1,25 @@
+
+# run this script from "07-make-figures.R"
+
+# --------------------- figure S6 ---------------------
+
+figure_s6 <- wq %>%
+  filter(param == "total aluminum", zone == 2, date < "2022-01-01") %>% 
+  mutate(
+    year = as.factor(year(date)),
+    yday = yday(date) / if_else(leap_year(date), 366, 365),
+    yday = 365 * yday + as.Date("2021-12-31")
+  ) %>% 
+  ggplot(aes(yday, value, col = year, group = interaction(series, year))) + 
+  scale_x_date(date_labels = "%b") +
+  scale_color_manual(values = palette[c(3,1,4,6)]) +
+  geom_point() +
+  geom_line() + 
+  ylim(c(0, 210)) +
+  labs(
+    x = NULL,
+    y = expression("[Al] (µg L"^-1*")"),
+    col = NULL
+  )
+
+ggsave("figures/figure-s6.png", figure_s6, dev = "png", dpi = 600, width = 6.5, height = 3)

### 2023-08-17 Rmarkdown/paper.Rmd
revision to text

@@
 
-Orthophosphate can control lead release into tap water, but its benefits are difficult to quantify since lead concentrations are so plumbing-dependent.
+Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific.
 Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified.
 The best sentinel homes, then, are often short-lived.
-Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at various locations within a water system.
+Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system.
 We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions.
-In this study, geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^.
+Geometric mean lead release from pipe racks exhibited a pronounced dose-response, falling by `r round(abs(pcent_decrease$y))`% after an increase from 1 to 2 mg PO~4~ L^-1^, and then climbing by `r round(abs(pcent_increase$y))`% after a decrease to 1.5 mg PO~4~ L^-1^.
 Data from nine sentinel homes were consistent with those from pipe racks: geometric mean lead at the high orthophosphate dose was `r d$d_50`% of that at the low dose.
@@
 While they overlap in form and function with pipe loops and bench apparatus, sentinel pipe racks are designed to estimate lead release from representative lead pipes into distributed drinking water with as much precision and accuracy as possible---in as close to real-time as possible.
-Sentinel pipe racks can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind.
+They can be used to understand the effect of an unplanned change in water quality, whereas pipe loop and bench-top studies are usually designed with a specific research question in mind.
 And while no simple model can fully replicate the complexities of premises plumbing [@devine_literature_2023], pipe rack systems are probably a better approximation than benchtop apparatus [@masters_comparison_2022].
@@
 
-A pronounced decreasing trend in lead concentration followed the increase to 2 mg PO~4~ L^-1^ (Figure 1a, c).
+A pronounced negative trend in lead concentration followed the increase to 2 mg PO~4~ L^-1^ (Figure 1a, c).
 Specifically, a 95% credible interval on the slope of the global multi-year trend excluded 0 µg Pb L^-1^ d^-1^ for a period beginning shortly after the dose increase and running until the end of the study period (d[Pb]/dt < 0, Figure 1a).
-The higher dose appeared to protect against lead release; and while this effect was slow to manifest, doubling the orthophosphate dose decreased the geometric mean lead concentration, year-over-year, by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
+The higher dose appeared to protect against lead release.
+And while this effect was slow to manifest, doubling the orthophosphate dose decreased the geometric mean lead concentration, year-over-year, by an estimated `r round(abs(pcent_decrease$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease, c(ymin, ymax)))))), collapse = "--")`%).
 
-Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding location-specific smooth term (Figure 1a).
+Lead release decreased faster in Zone 2 than in Zone 1---an effect that was apparent in the corresponding Zone 2-specific smooth term (Figure 1a).
 One explanation is a modified treatment process: coagulation pH at the Zone 2 treatment plant was increased from less than 6 to approximately 6.3 in April 2021 (Figure S4b), to target the pH of minimum aluminum hydroxide solubility [@edzwald_aluminum_2020].
@@
 Furthermore, a 95% credible interval on the slope of the Zone 2 trend excluded 0 µg Pb L^-1^ d^-1^ for several months, beginning shortly after the pH increase (d[Pb]/dt < 0, Figure 1b).
-Changes to the coagulation process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%).
+Changes to the treatment process, then, appear to have lowered lead release: between the coagulation pH increase and the orthophosphate dose increase, geometric mean lead decreased by an estimated `r round(abs(pcent_decrease2$y))`% (95% credible interval: `r paste(suppressWarnings(sort(round(abs(with(pcent_decrease2, c(ymin, ymax)))))), collapse = "--")`%).
 Since only a short period separated the pH increase and the change in orthophosphate dose, controlling for orthophosphate's effect yielded a more reliable estimate of the effect we attributed to a change in coagulation pH.
@@
 
-Even without an informative prior on $\beta$ (equation 2), the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks.
+Even without an informative prior on $\beta$ (equation 2), however, the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks.
 This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016].
-And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we should expect lower variability [@masters_inherent_2016].
+And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we expect lower variability [@masters_inherent_2016].
 
@@
 A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b).
-We used the sentinel pipe racks to determine the orthophosphate dose response in this zone.
-We estimated the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and the effect of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
+We used the sentinel pipe racks to determine the orthophosphate dose response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
 But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016].
-Seasonality was more complex in Zone 2, perhaps due to the inverse seasonality in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]).
-And since the dose increases occurred in late November and July in Zones 1 and 2 respectively, controlling explicitly for seasonality in the full model---encompassing both zones---was less important.
+Seasonality was more complex in Zone 2, perhaps due to the inverse seasonal pattern in aluminum (especially before the change in coagulation pH [@trueman_seasonal_2022]).
+And because the dose increase effect was estimated year-over-year in the full model (encompassing both zones) controlling for seasonality was less important.
 
 Temperature variation in the pipe rack effluent was a reasonable match for distributed water.
-While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked those of distributed water and point-of-use flushed samples more closely than the above-ground site (Rack 2), both captured the seasonal pattern well.
+While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked distributed water and point-of-use flushed sample temperatures more closely than the above-ground site (Rack 2), both captured the seasonal pattern well.
+This, of course, is not necessarily true---in general, a comparison of pipe rack effluent and distributed water temperatures will help to illustrate the limitations of the pipe rack data.
 
@@
 
-Even after accounting for the seasonal variation in lead release, though, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a) and a 95% credible interval on the slope of the global trend that excluded zero.
+And even after accounting for the seasonal variation in lead release, a decrease in the orthophosphate dose to 1.5 mg PO~4~ L^-1^ was followed by an increase in lead release (Figure 3a)---and a 95% credible interval on the slope of the global trend that excluded zero.
 The intermediate dose appears to have yielded lead concentrations between those resulting from the 1 and 2 mg PO~4~ L^-1^ doses.

### 2023-08-18 Rmarkdown/paper.Rmd
minor edits to manuscript text

@@
 ---
-
-<!-- to do: -->
-<!-- 1. cite https://doi.org/10.1080/00031305.2018.1549100 for bayesian r2? -->
 
@@
 The best sentinel homes, then, are often short-lived.
-Here, we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system.
+Here we explore an alternative: recovered lead pipe racks supplied with distributed drinking water at locations throughout a water system.
 We also propose a strategy for analyzing the data based on the generalized additive model, which approximates time series as sums of smooth functions.
@@
 Water quality from both sources is well suited to orthophosphate corrosion control treatment [@schock1996corr; @trueman_effects_2018], with median pH and dissolved inorganic carbon concentrations in pipe rack effluent of 7.3--7.4 and 4--5 mg C L^-1^ (Table 1).
-But while water quality in Zones 1 and 2 was largely similar, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022].
+But while most water quality parameters were similar across zones, aluminum concentrations were markedly different: aluminum in Zone 2 was seasonal, with peak concentrations occurring at the coldest water temperatures [@trueman_seasonal_2022].
 Aluminum concentrations in Zone 1 were much lower and more consistent throughout the year (Table S1).
@@
 Doubling the orthophosphate dose decreased lead release from pipe racks substantially.
-At 1 mg PO~4~ L^-1^, lead in pipe rack effluent did not exhibit a clear time-trend  (Figure 1c).
+At the low dose of 1 mg PO~4~ L^-1^, lead in pipe rack effluent did not exhibit a clear time-trend (Figure 1c)---an indication that the pipe racks had stabilized.
 In particular, a 95% credible interval on the slope of the global multi-year trend---capturing variation common to all pipe sections---included 0 µg Pb L^-1^ d^-1^ everywhere (d[Pb]/dt ~ 0, Figure 1a).
-Pipe racks, then, appear to have stabilized well at the initial orthophosphate concentration.
 
@@
 Moreover, an improved coagulation process might remove more of the natural organic matter fractions that increase lead solubility by complexation [@korshin_corrosion_1999].
-But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data are not rich enough to identify the true cause.
+But while we would expect lower aluminum, higher orthophosphate, and better organic matter removal all to decrease lead release, the data we have are not sufficient to explain the trend conclusively.
 
@@
 
-Temperature variation in the pipe rack effluent was a reasonable match for distributed water.
+Temperature in the pipe rack effluent was a reasonable match for distributed water temperature.
 While the underground site---Rack 1 (Figure 3e)---yielded effluent water temperatures that tracked distributed water and point-of-use flushed sample temperatures more closely than the above-ground site (Rack 2), both captured the seasonal pattern well.

### 2023-08-21 Rmarkdown/paper.Rmd
chemrxiv resubmission

@@
 Orthophosphate can limit lead contamination of tap water, but its benefits are difficult to quantify since lead concentrations are so site-specific.
-Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate removal of lead once identified.
+Sentinel homes serviced by lead pipe are ideal for monitoring orthophosphate treatment, but best practices dictate the removal of lead once identified.
 The best sentinel homes, then, are often short-lived.
@@
 Updated regulations on lead in drinking water promise to expedite replacement of lead service lines in Canada and the USA.
-Even afterwards, though, a substantial legacy of lead plumbing---including lead:tin solder and brass---will have to be managed.
+Even afterward, though, a substantial legacy of lead plumbing---including lead:tin solder and brass---will have to be managed.
 This will require careful control of drinking water chemistry to limit lead solubility and maintain durable corrosion scale.
@@
 Then, we used this estimate as a prior probability for the same effect in nine sentinel homes.
-Finally, we quantified the orthophosphate dose response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^.
+Finally, we quantified the orthophosphate dose-response of a subset of the pipe racks at 1, 2, and then 1.5 mg PO~4~ L^-1^.
 Our results provide a Bayesian framework for analyzing corrosion control treatment data, especially when they are collected as time series and have multiple sources.
@@
   ) %>% 
-  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent by zone, with the exception of hardness, which was measured in distributed water samples.")
+  knitr::kable(caption = "**Table 1.** Summary of water quality in pipe rack effluent by zone, except for hardness, which was measured in distributed water samples.")
 ```
@@
 All samples were collected in wide-mouth plastic bottles and acidified with trace metal-grade nitric acid.
-Volunteers were instructed to open their faucets all the way and not to adjust the flow rate while sampling (Figure S3).
+Volunteers were instructed to open their faucets all the way and not to adjust the flow rate while sampling, but they were not instructed to measure flow rate (Figure S3).
 
@@
 Even without an informative prior on $\beta$ (equation 2), however, the point-of-use data yielded a more precise estimate of the effect of doubling the orthophosphate dose, with a 95% credible interval that was approximately half as wide as the one estimated from the pipe racks.
-This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically-coupled at both ends to brass, we expect high variability [@masters_inherent_2016].
+This is mainly a result of the differences in materials and lead levels; since the pipe racks represent lead release from lead pipe galvanically coupled at both ends to brass, we expect high variability [@masters_inherent_2016].
 And since the point-of-use data more probably represent lead release from a variety of sources---and at a much lower level---we expect lower variability [@masters_inherent_2016].
@@
 A little more than a year after the orthophosphate dose was increased in Zone 1, it was decreased from 2 to 1.5 mg PO~4~ L^-1^ (Figure 3b).
-We used the sentinel pipe racks to determine the orthophosphate dose response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
+We used the sentinel pipe racks to determine the orthophosphate dose-response in this zone, by estimating the effect of an increase from 1 to 2 mg PO~4~ L^-1^ and of a subsequent decrease to 1.5 mg PO~4~ L^-1^.
 But since the final decrease occurred in the spring---as water temperatures were increasing rapidly (Figure 3e)---we estimated the seasonal variation in lead release and added it as a separate term in the model to control for temperature effects [@masters_seasonal_2016].
@@
 Passivation/maintenance dosing has the potential to conserve phosphorus, but it should be evaluated carefully to avoid unwanted increases in lead release at the maintenance dose or excess particulate lead at an unnecessarily high passivation dose [@zhao_formation_2018; @trueman_comparing_2022].
-Here though, the dose response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.
+Here though, the dose-response of lead release to orthophosphate was qualitatively similar to that predicted by solubility: lead release decreased when orthophosphate was increased and increased when orthophosphate was decreased.
 
@@
 But lead service line replacement, incomplete participation by residents in sampling programs, and changes to premises plumbing make it difficult to monitor the effectiveness of corrosion control over time this way.
-And while no simple apparatus can reliable quantify human exposure to lead, sentinel pipe racks offer an alternative to point-of-use sampling for non-regulatory monitoring.
+And while no simple apparatus can reliably quantify human exposure to lead, sentinel pipe racks offer an alternative to point-of-use sampling for non-regulatory monitoring.
 Especially when installed at multiple locations across a water distribution network, sentinel pipe racks can be used to understand how both anticipated and unexpected changes in water quality impact lead concentrations.

### 2023-08-28 README.Rmd

@@
 
-This repository contains the code and data necessary to reproduce the main results of the preprint ["Sentinel lead pipe racks quantify orthophosphate’s dose-response in drinking water"](https://doi.org/10.26434/chemrxiv-2023-h1459).
+This repository contains the code and data necessary to reproduce the main results of the preprint ["Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control"](https://doi.org/10.26434/chemrxiv-2023-h1459).

### 2023-10-16 README.Rmd
update readme with link to published paper

--- 2023-05-26 README.Rmd
@@
 
-This repository contains the code and data necessary to reproduce the main results of the preprint ["Sentinel lead pipe racks quantify orthophosphate’s dose-response in drinking water"](https://doi.org/10.26434/chemrxiv-2023-h1459).
+This repository contains the code and data necessary to reproduce the main results of the paper ["Evaluating sentinel pipe racks for monitoring lead release and optimizing corrosion control"](https://doi.org/10.1021/acsestwater.3c00273) (also available as a [preprint](https://doi.org/10.26434/chemrxiv-2023-h1459-v3)).
